### 2024-11-29 _paper/paper.qmd
work-in-progress on paper

---
title: Modelling potential environmental impacts of science activity in Antarctica
author: 
  - name: David O’Sullivan
    email: david.osullivan@auckland.ac.nz
    affiliations:
      - name: University of Auckland
        url: https://profiles.auckland.ac.nz/david-osullivan
      - name: Geospatial Stuff 
        url: https://dosull.github.io
  - name: Fraser J. Morgan
    affiliations: 
      - name: Manaaki Whenua Landcare Research 
        url: https://landcareresearch.co.nz
bibliography: antarctica.bib
csl: the-canadian-geographer.csl
format: 
  html:
    toc: true
    number-sections: true
    number-depth: 3
    self-contained: true
  pdf:
    documentclass: article
    classoption: [11pt]
    toc: false
    number-sections: true
    number-depth: 3
execute:
  echo: false
  warning: false
  results: false
  message: false
---

### Abstract {.unnumbered}
We use GPS data collected on a science expedition in Antarctica to estimate hiking functions for the speed at which humans traverse terrain differentiated by slope and by ground cover (moraines and rock). We use the estimated hiking functions to build weighted directed graphs as a representation of specific environments in Antarctica. From these we estimate using a variety of graph metrics—particularly betweennness centrality—the relative potential for human environmental impacts arising from scientific activities in those environments. We also suggest a simple approach to planning science expeditions that might allow for reduced impacts in these environments.

## Introduction
Overview of Antarctic science since when and where, its intensity etc. Background on international treaties, etc.

Relevant findings as to human impacts in Antarctica. 
Note that in this environment even ‘leave only footprints’ is likely impacting the environment in significant ways.

Overview of sections ahead.

## Our approach and related work
We chose to explore the question of where human impacts are likely to be strongest using an approach closely related to work on patterns of human movement in archaeology [@Verhagen2019] where likely and potential movement paths of humans across landscapes have been used to infer the settlement structure and human geography of large-scale landscapes. 
Closely related work in biology investigates the structure and geography of animal transportation networks [@Perna2014]. 
Both approaches rely on the idea that humans or animals move around in an environment in time or energy efficient ways. 
These approaches rely on hiking functions that relate speed of movement across a terrain to its slope.

Hiking functions must be applied in some context where locations across a landscape are connected to one another. 
Because hiking functions are asymmetric, with estimated speed of movement up slopes different than estimated speeds down the same slope, landscape must be represented in a way that allows for this asymmetry. 
We therefore represent terrain in our landscapes as directed graphs (or network) of connections between locations regularly distributed in planimetric space across the landscape of interest. 
Because the graphs are directed the costs associated with movement between two locations can be different depending on the direction of movement. 
Additionally, we associate with locations (i.e., vertices in the graph) the ground cover at the location, which also affects the speed at which it can be traversed. 
Because the ground cover in the Antarctic environments under study can be broadly categorised into only two types, moraine and rock, we use the ground cover of a location to switch between two estimated hiking functions, rather than the more widely used approach of penalising movement on different ground covers by applying cost factors. 
We consider previous work on hiking functions and directed graphs in more detail below.

### Hiking functions
@Prisner2023 provide an overview of a variety of functions that have been used to model how hiking times and speeds vary with terrain slope. 
They consider longstanding rules of thumb [@Naismith1892], and later modifications thereof [@Langmuir1984], along with more recent such guidance from the Swiss and German Alpine Clubs [@Winkler2010;@Deutscher2011]. 
These functions estimate the time taken to travel 1km, referred to as _pace_, based on slope expressed as _rise over run_, that is change in elevation divided by horizontal distance. 
They are all piecewise functions with sharp changes in estimated pace at specific slopes.

Alongside these hiking pace functions @Prisner2023 also present hiking speed functions (generally referred to as simply _hiking functions_) from Tobler [-@Tobler1993 generally considered the first hiking function] and more recent, related but more firmly empirically grounded alternatives offered by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019. Another hiking function not discussed by @Prisner2023 is presented by @Rees2004. 
These hiking functions are all continuous in the slope of the terrain so that $v=f(\theta)$, where $v$ is the speed, and $\theta$ is the slope. 
They can all be parameterised to control the maximum speed attainable, the slope at which maximum speed is attained (expected to be a shallow downhill slope), and the rate at which speed falls off with increasing slope.

The functional form of some of these functions is shown in @tbl-hiking-function-equations and graphed in @fig-hiking-functions. 
The parameterisation of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the Exponential hiking function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
As has been noted elsewhere [see @Herzog2010;@Campbell2019] these parameter settings for the Tobler function are based on a loose fit with secondary data on hiking speeds presented in Imhof [-@Imhof1950, pp. 217-220], although these exact values are frequently applied in the literature.

| Description | Equation   | Examples
|:------------|------------|:--------
|Exponential  | $ae^{b\left|\theta-c\right|}$ | @Tobler1993, @Marquez-Perez2017
| Gaussian    | $ae^{-b(\theta-c)^2}$ | @Irmischer2018, @Campbell2019
| Lorentz     | $\frac{a}{\left[b+d(\theta-c)^2\right]}$ | @Campbell2019
| Quadratic   | $a+b\theta+c\theta^2$ | @Rees2004

: Functional forms of hiking functions {#tbl-hiking-function-equations .striped tbl-colwidths="[20,20,80]"}

```{r}
#| label: fig-hiking-functions
#| fig-cap: Hiking functions.

library(ggplot2)
library(cols4all)
library(tidyr)
library(dplyr)

theme_set(theme_minimal())

tobler <- function(theta, a = 6, b = 3.5, c = -0.05) {
  a * exp(-b * abs(theta - c))
}
gaussian <- function(theta, a = 5.5, b = 0.35, c = -0.04) {
  a * exp(-(theta - c) ^ 2 / b ^ 2)
}

lorentz <- function(theta, a = 5.25, b = 1, c = -0.06, d = 10) {
  a / (b + d * (theta - c) ^ 2)
}

rees <- function(theta, a = 5.25, b = -1.5, c = -16) {
  5 + b * theta + c * theta ^ 2
}

results <- tibble(Slope = -50:50/100,
                  Tobler = tobler(Slope), 
                  Gaussian = gaussian(Slope),
                  Lorentz = lorentz(Slope),
                  Rees = rees(Slope)) |>
  pivot_longer(cols = -1) |>
  rename(Function = name, `Speed, km/h` = value)

ggplot(results) +
  geom_line(aes(x = Slope, y = `Speed, km/h`, colour = Function, group = Function)) +
  scale_colour_discrete_c4a_cat("brewer.set1")
```

An additional example is offered by @Wood2023 who include the local gradient of the terrain (i.e. the maximum slope at each location, not the slope in the direction of movement) as a parameter in the function.

There is no generally accepted preferred form for hiking functions, although Tobler’s function has been the most widely used.

Note the strange origin of the popular exponential Tobler function (in a working paper, based on data in a figure in a relatively obscure Swiss book from 1950, and... according to Herzog 2010 a very poor fit to those data in any case...)

Edge costs in graphs modelled based on slope and a hiking function

### Representing the landscape
Terrain is asymmetric with respect to movement, therefore: landscapes as graphs

Movement cost for graph edges derived from hiking functions

Graph vertices augmented with landcover information (in this context limited to surface geology)

## Data sources
### Antarctic geospatial data
@Cox2023, @Cox2023a, @Felden2023

[FIGURE showing location information etc., based on these data]

### GPS data from an expedition

Devices used, and associated protocols for scientists while on site.

[FIGURE showing GPS traces in context]

Initial filtering to remove implausibly high-speed movement fixes – these largely related to helicopter movements and arrival/departure from the valley.

Additional cleanup of GPS data required: filtering to remove many GPS fixes with associated low movement rates associated with time spent at base camps, rest stops, or experimental sites.

Crude filters: short distances between fixes, large turn angles between fixes, and seemingly anomalous high speeds recorded at 0 slope angles (yet to be explained but required to obtain reasonable results).

Density-based filtering: hexbinning of fixes to identify non-purposive movement areas.

[FIGURE showing filtering steps and impact on slope vs speed scatterplot/boxplots]

## Methods and results 
### Hiking functions
Fit three alternative functional forms: exponential Tobler (1993, pp 1–4, based on data from Imhof 1950), Gaussian Irmischer and Clarke (2018), and Student’s t distribution, the last of these as a heavier-tailed alternative to Gaussian. Other approaches rely on rule-of-thumb hiking time calculations such as...

The Gaussian form offered the best overall fits to our data. Although Tobler’s exponential form is much used it is somewhat implausible that hiking speeds would peak sharply at a very specific slope angle, and so we adopt the Gaussian form.

[FIGURE showing hiking function fits about here]

In previous work researchers have applied a ground cover penalty cost to a base hiking function to estimate traversal times. In our case there are effectively only two terrain types, moraines (gravel) and bare rock. This makes the option of fitting different functions to each attractive and reasonable, and the results of fitting distinct functions to each support this approach. The peak speed on bare rock is attained on steeper downhill slopes than on gravel, perhaps indicative of the greater care required on downhill gravel slopes. Meanwhile the highest speeds on level terrain are attained on gravel. Our fitted functions are shown below.

[FIGURE comparing moraine and rock hiking functions about here]

### Landscapes as graphs
Assignment of nodes as a lattice of equally spaced locations. Choice of rectangular/square grid vs. hexagonal. Square grids are easily generated and also offer some computational advantages if, for example, the grid is aligned with an underlying digital elevation model (DEM). However, square grids present the problem of choosing between two different potential local neighbourhoods, based on orthogonal adjacency only, or on orthogonal and diagonal adjacency (respectively referred to as rook and queen’s case adjacency, or as von Neumann and Moore neighbourhoods, in the spatial analysis and cellular automaton literatures). Excluding diagonal adjacency yields implausible diamond shaped isochrones on level terrain. Including diagonal adjacency can lead to seemingly contradictory results when alternate corners of a square are at different heights.

Given these challenges we adopt a hexagonally arranged triangular lattice. This avoids any difficulties with varying distances (i.e. network edge lengths) between nodes. Note that this is necessarily an approximation, and on extensive flat terrains will yield unrealistic hexagonal isochrones. It also presents some computational disadvantages in assigning elevations to graph nodes where interpolation of heights is required due to mismatches between hexagon centres and cells in grid-based DEMs.

[FIGURE showing square and hexagonal lattices and issues]

Alternative approaches might estimate movement based on generating multiple random approximately evenly spaced nodes across a terrain and combining the resulting estimated travel times [see @Etherington2012], or deploying methods for estimating geodesics on complex triangular meshes [@Martinez2005] but both approaches are more computationally demanding than the current application justifies. In any case, the latter method does not allow for asymmetry in traversal times as arising due to slopes.

We developed R code [@RCoreTeam2024] to build graphs (i.e. networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions. Graphs are stored as igraph package [@csardi2006igraph;@Csardi2024] graph objects for further analysis.

How terrain is handled in graph edge cost estimation (i.e., assigning half the cost from the terrain at the vertex at each end of the edge).

`igraph` implements betweenness centrality measures [@Freeman1978;@Brandes2001], which we used to identify vertices and/or edges most likely to be traversed by minimum travel time routes across the terrain.

### Betweenness centrality limited by radius.

### Impact minimizing networks
Tentative proposal for impact minimizing networks based on minimum spanning trees, but noting the issue with respect to directed graphs when these would more correctly be arborescences [@Korte2018a].

## Discussion
Blah

## Conclusions
Blah

### 2024-11-29 paper/paper.qmd
put paper on website

--- 2024-11-29 _paper/paper.qmd
@@
 ## Introduction
-Overview of Antarctic science since when and where, its intensity etc. Background on international treaties, etc.
+Overview of Antarctic science: when and where, its intensity etc. Background on international treaties, etc.
 
@@
 The functional form of some of these functions is shown in @tbl-hiking-function-equations and graphed in @fig-hiking-functions.
-The parameterisation of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the Exponential hiking function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
-As has been noted elsewhere [see @Herzog2010;@Campbell2019] these parameter settings for the Tobler function are based on a loose fit with secondary data on hiking speeds presented in Imhof [-@Imhof1950, pp.
-217-220], although these exact values are frequently applied in the literature.
 
@@
 #| label: fig-hiking-functions
-#| fig-cap: Hiking functions.
+#| fig-cap: Example hiking functions.
 
@@
 results <- tibble(Slope = -50:50/100,
-                  Tobler = tobler(Slope), 
+                  Exponential = tobler(Slope), 
                   Gaussian = gaussian(Slope),
@@
 ggplot(results) +
-  geom_line(aes(x = Slope, y = `Speed, km/h`, colour = Function, group = Function)) +
-  scale_colour_discrete_c4a_cat("brewer.set1")
+  geom_line(aes(x = Slope, y = `Speed, km/h`, 
+            colour = Function, group = Function), linewidth = 1) +
+  scale_colour_discrete_c4a_cat("brewer.set2") +
+  scale_x_continuous(limits = c(-0.5, 0.5), expand = c(0, 0)) +
+  theme(panel.border = element_rect(linewidth = 0.5, fill = NA))
 ```
 
-An additional example is offered by @Wood2023 who include the local gradient of the terrain (i.e.
-the maximum slope at each location, not the slope in the direction of movement) as a parameter in the function.
+The parameterisation of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the Exponential hiking function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
+These parameter values for the exponential functional form give _Tobler's hiking function_ but, as has been noted elsewhere, [see @Herzog2010;@Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds presented by Imhof [-@Imhof1950, pp.
+217-220].
+Nevertheless these parameter values are widely applied in the literature.
 
-There is no generally accepted preferred form for hiking functions, although Tobler’s function has been the most widely used.
+All these hiking function forms are somewhat _ad hoc_. 
+They exhibit desirable, expected properties: a peak speed at a slope near zero, which we expect to be slightly negative (i.e., downhill), and continuously falling speeds as the slope deviates from the slope of peak speed.
+However, there is no theoretical basis for the specific functional forms listed in @tbl-hiking-function-equations.
+More principled approaches might be developed based on the literature on the physiology of human movement, see e.g. **CITATIONS NEEDED HERE**.
+In general, approaches based on minimising energy use yield similar results to empirical speed-slope functions, although it is worth noting that they more reliably generate zig-zag or 'switchback' movement behaviour on steep slopes [@Llobera2007].
+However, these approaches are hard to implement, and empirically-derived and locally-specific hiking functions as reported by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019 are the approach we have taken in the present work, based on available data and the goals of our study where the relative cost of different potential routes in the landscape is more important than exact prediction of routes.
+In practice, almost any function with the peaked form of those shown in @fig-hiking-functions is suited to our requirements.
 
-Note the strange origin of the popular exponential Tobler function (in a working paper, based on data in a figure in a relatively obscure Swiss book from 1950, and...
-according to Herzog 2010 a very poor fit to those data in any case...)
-
-Edge costs in graphs modelled based on slope and a hiking function
+It is commonplace in many applications to also incorporate a penalty on movement contingent on landcover, especially for off-track or off-road movement.
+For example the speed attainable off-track in forested terrain might be only half that attainable in grasslands.
+Unsurprisingly, there are no widely agreed landcover penalties, but see for example, those compiled by @Herzog2020.
+The hiking function reported by @Wood2023 includes the local gradient of the terrain (i.e., the maximum slope at each location, not the slope in the direction of movement) as a covariate in the estimated function.
+It is possible that this kind of approach treating landcover as a categorical covariate in the estimation of more complex hiking functions might be applied in future work. 
+In our application because there are only two kinds of navigable landcover&mdash;moraine (or gravel) and rock&mdash;we chose instead to estimate two hiking functions, one for each landcover and estimate movement costs conditional on the landcover present at each location.
+This has the additional advantage of allowing for distinctly different effects of slope on speed.
+Details of this approach are reported and discussed in @sec-hiking-function-estimation.
 
@@
 Terrain is asymmetric with respect to movement, therefore: landscapes as graphs
+
+Edge costs in graphs modelled based on slope and a hiking function
 
@@
 ## Methods and results 
-### Hiking functions
+### Hiking functions {#sec-hiking-function-estimation}
 Fit three alternative functional forms: exponential Tobler (1993, pp 1–4, based on data from Imhof 1950), Gaussian Irmischer and Clarke (2018), and Student’s t distribution, the last of these as a heavier-tailed alternative to Gaussian.

### 2024-12-01 paper/paper.qmd
latest changes to paper

@@
 csl: the-canadian-geographer.csl
+lightbox: true
 format: 
@@
     number-depth: 3
+    linestretch: 1.15
 execute:
@@
 ---
+```{css, echo = FALSE}
+p {color: black;}
+figcaption {color: black;}
+```
+
+```{r}
+library(ggplot2)
+library(ggpubr)
+library(grid)
+library(cols4all)
+library(tidyr)
+library(dplyr)
+library(igraph)
+library(sf)
+```
 
 ### Abstract {.unnumbered}
+
 We use GPS data collected on a science expedition in Antarctica to estimate hiking functions for the speed at which humans traverse terrain differentiated by slope and by ground cover (moraines and rock).
@@
 ## Introduction
+
 Overview of Antarctic science: when and where, its intensity etc. Background on international treaties, etc.
 
-Relevant findings as to human impacts in Antarctica. 
-Note that in this environment even ‘leave only footprints’ is likely impacting the environment in significant ways.
+Relevant findings as to human impacts in Antarctica. Note that in this environment even ‘leave only footprints’ is likely impacting the environment in significant ways.
 
@@
 ## Our approach and related work
+
 We chose to explore the question of where human impacts are likely to be strongest using an approach closely related to work on patterns of human movement in archaeology [@Verhagen2019] where likely and potential movement paths of humans across landscapes have been used to infer the settlement structure and human geography of large-scale landscapes.
@@
 ### Hiking functions
+
 @Prisner2023 provide an overview of a variety of functions that have been used to model how hiking times and speeds vary with terrain slope. 
-They consider longstanding rules of thumb [@Naismith1892], and later modifications thereof [@Langmuir1984], along with more recent such guidance from the Swiss and German Alpine Clubs [@Winkler2010;@Deutscher2011].
-These functions estimate the time taken to travel 1km, referred to as _pace_, based on slope expressed as _rise over run_, that is change in elevation divided by horizontal distance. 
+They consider longstanding rules of thumb [@Naismith1892], and later modifications thereof [@Langmuir1984], along with more recent such guidance from the Swiss and German Alpine Clubs [@Winkler2010; @Deutscher2011].
+These functions estimate the time taken to travel 1km, referred to as *pace*, based on slope expressed as *rise over run*, that is change in elevation divided by horizontal distance. 
 They are all piecewise functions with sharp changes in estimated pace at specific slopes.
 
-Alongside these hiking pace functions @Prisner2023 also present hiking speed functions (generally referred to as simply _hiking functions_) from Tobler [-@Tobler1993 generally considered the first hiking function] and more recent, related but more firmly empirically grounded alternatives offered by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019.
+Alongside these hiking pace functions @Prisner2023 also present hiking speed functions (generally referred to as simply *hiking functions*) from Tobler [-@Tobler1993 generally considered the first hiking function] and more recent, related but more firmly empirically grounded alternatives offered by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019.
 Another hiking function not discussed by @Prisner2023 is presented by @Rees2004. 
@@
 
-| Description | Equation   | Examples
-|:------------|------------|:--------
-|Exponential  | $ae^{b\left|\theta-c\right|}$ | @Tobler1993, @Marquez-Perez2017
-| Gaussian    | $ae^{-b(\theta-c)^2}$ | @Irmischer2018, @Campbell2019
-| Lorentz     | $\frac{a}{\left[b+d(\theta-c)^2\right]}$ | @Campbell2019
-| Quadratic   | $a+b\theta+c\theta^2$ | @Rees2004
+| Description | Equation                                 | Examples                        |
+|:--------------------------|-------------------------|:------------------|
+| Exponential | $ae^{b\left|\theta-c\right|}$            | @Tobler1993, @Marquez-Perez2017 |
+| Gaussian    | $ae^{-b(\theta-c)^2}$                    | @Irmischer2018, @Campbell2019   |
+| Lorentz     | $\frac{a}{\left[b+d(\theta-c)^2\right]}$ | @Campbell2019                   |
+| Quadratic   | $a+b\theta+c\theta^2$                    | @Rees2004                       |
 
-: Functional forms of hiking functions {#tbl-hiking-function-equations .striped tbl-colwidths="[20,20,80]"}
+: Functional forms of hiking functions {#tbl-hiking-function-equations .striped tbl-colwidths="\[20,20,80\]"}
 
@@
 #| label: fig-hiking-functions
-#| fig-cap: Example hiking functions.
-
-library(ggplot2)
-library(cols4all)
-library(tidyr)
-library(dplyr)
+#| fig-cap: "Example hiking functional forms: Exponential [@Tobler1993], Gaussian [@Irmischer2018], Lorentz [@Campbell2019], and Quadratic [@Rees2004]."
 
 theme_set(theme_minimal())
+theme_update(panel.border = element_rect(linewidth = 0.5, fill = NA))
 
@@
 }
-gaussian <- function(theta, a = 5.5, b = 0.35, c = -0.04) {
-  a * exp(-(theta - c) ^ 2 / b ^ 2)
+gaussian <- function(theta, a = 5.5, b = 0.25, c = -0.04) {
+  a * exp(-(theta - c) ^ 2 / 2 / b ^ 2)
 }
-
-lorentz <- function(theta, a = 5.25, b = 1, c = -0.06, d = 10) {
-  a / (b + d * (theta - c) ^ 2)
+lorentz <- function(theta, a = 4.75, b = 0.9, c = -0.06, d = b / pi) {
+  a / (b * (1 + ((theta - c) / d) ^ 2))
 }
-
-rees <- function(theta, a = 5.25, b = -1.5, c = -16) {
+rees <- function(theta, a = 5.25, b = -1.5, c = -16.5) {
   5 + b * theta + c * theta ^ 2
@@
                   Lorentz = lorentz(Slope),
-                  Rees = rees(Slope)) |>
+                  Quadratic = rees(Slope)) |>
   pivot_longer(cols = -1) |>
-  rename(Function = name, `Speed, km/h` = value)
+  rename(`Functional form` = name, `Speed, km/h` = value)
 
 ggplot(results) +
-  geom_line(aes(x = Slope, y = `Speed, km/h`, 
-            colour = Function, group = Function), linewidth = 1) +
-  scale_colour_discrete_c4a_cat("brewer.set2") +
+  geom_line(aes(x = Slope, y = `Speed, km/h`, linetype = `Functional form`,
+            colour = `Functional form`, group = `Functional form`), linewidth = 1) +
+  scale_colour_discrete_c4a_cat("brewer.set1") +
+  scale_linetype_manual(values = c("solid", "longdash", "dotdash", "dotted")) + 
   scale_x_continuous(limits = c(-0.5, 0.5), expand = c(0, 0)) +
-  theme(panel.border = element_rect(linewidth = 0.5, fill = NA))
+  scale_y_continuous(limits = c(0, 6.25), expand = c(0, 0)) +
+  geom_vline(aes(xintercept = -0.05), lwd = 0.35, linetype = "dotted") +
+  annotate("label", x = -0.055, y = 0.5, label = "Peak speed slope (exponential)", 
+           angle = 90, hjust = 0, vjust = 0.6, label.size = 0)
 ```
 
-The parameterisation of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the Exponential hiking function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
-These parameter values for the exponential functional form give _Tobler's hiking function_ but, as has been noted elsewhere, [see @Herzog2010;@Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds presented by Imhof [-@Imhof1950, pp.
+The parameterisations of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the exponential function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
+These parameters give *Tobler's hiking function* but, as has been noted elsewhere, [see @Herzog2010; @Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds presented by Imhof [-@Imhof1950, pp.
 217-220].
@@
 
-All these hiking function forms are somewhat _ad hoc_. 
+All these hiking functional forms are somewhat *ad hoc*. 
 They exhibit desirable, expected properties: a peak speed at a slope near zero, which we expect to be slightly negative (i.e., downhill), and continuously falling speeds as the slope deviates from the slope of peak speed.
@@
 In general, approaches based on minimising energy use yield similar results to empirical speed-slope functions, although it is worth noting that they more reliably generate zig-zag or 'switchback' movement behaviour on steep slopes [@Llobera2007].
-However, these approaches are hard to implement, and empirically-derived and locally-specific hiking functions as reported by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019 are the approach we have taken in the present work, based on available data and the goals of our study where the relative cost of different potential routes in the landscape is more important than exact prediction of routes.
+However, these are hard to implement, and we have adopted empirically-derived and locally-specific hiking functions following @Marquez-Perez2017, @Irmischer2018, and @Campbell2019. 
+This choice is based on available data and the goals of our study, where the relative cost of different potential routes in the landscape is more important than exact prediction of routes. 
 In practice, almost any function with the peaked form of those shown in @fig-hiking-functions is suited to our requirements.
@@
 Unsurprisingly, there are no widely agreed landcover penalties, but see for example, those compiled by @Herzog2020. 
-The hiking function reported by @Wood2023 includes the local gradient of the terrain (i.e., the maximum slope at each location, not the slope in the direction of movement) as a covariate in the estimated function.
-It is possible that this kind of approach treating landcover as a categorical covariate in the estimation of more complex hiking functions might be applied in future work. 
-In our application because there are only two kinds of navigable landcover&mdash;moraine (or gravel) and rock&mdash;we chose instead to estimate two hiking functions, one for each landcover and estimate movement costs conditional on the landcover present at each location.
-This has the additional advantage of allowing for distinctly different effects of slope on speed.
+A hiking function derived by @Wood2023 includes the local gradient of the terrain (i.e., the maximum slope at each location, not the slope in the direction of movement) as a covariate in the estimated function.
+It is possible that this kind of approach treating landcover as a categorical covariate in the estimation of more complex hiking functions might be more widely applied in future work. 
+In our application, because there are only two kinds of navigable landcover&mdash;moraine (or gravel) and rock&mdash;we chose instead to estimate two hiking functions, one for each landcover, and estimate movement costs conditional on the landcover present at each location.
+This also has the advantage of allowing for differing effects of slope on speed due to landcover, where for example, gravel might allow more rapid movement on the level surfaces, but more rapidly reduce speed on slopes.
 Details of this approach are reported and discussed in @sec-hiking-function-estimation.
@@
 ### Representing the landscape
-Terrain is asymmetric with respect to movement, therefore: landscapes as graphs
+Hiking functions are usually asymmetric with respect to slope, with the highest attainable speed at a slight downhill slope. 
+This asymmetry means that unless analysis is focused on assessing movement cost from a single origin or to a single destination, it is necessary to represent the landscape in a way that can accommodate asymmetry.
+We therefore represent the landscape as a directed graph. 
+Graph vertices are locations with associated elevation and landcover. 
+Graph edges are directed links between vertices for which a change in elevation between the start and end vertex can be calculated, and a slope derived, based on the elevation difference and horizontal distance between the vertices.
 
-Edge costs in graphs modelled based on slope and a hiking function
+The most obvious way to derive such a graph is to assign each cell in a raster digital elevation model (DEM) to a graph vertex, so that graph vertices are arranged in a regular square grid or lattice.
+This approach has been widely adopted **CITATIONS NEEDED HERE** 
+However, this still requires a decision to be made regarding the definition of graph edges, which can be thought of as _allowed moves_ between vertices. 
+Three possibilities on a square lattice&mdash;Rook's, Queen's, and Knight's moves&mdash;are shown in @fig-lattice-adjacencies.
+Another option shown in the figure is to lay out a regular hexagonal grid of vertex locations and link each vertex with its six nearest neighbours.
 
-Movement cost for graph edges derived from hiking functions
+```{r}
+#| label: fig-lattice-adjacencies
+#| fig-cap: Possible graph lattices.
+#| fig-width: 7
+#| fig-height: 7
 
-Graph vertices augmented with landcover information (in this context limited to surface geology)
+theme_set(theme_void()) 
+theme_update(plot.margin = unit(c(1, 1, 2, 1), "mm"),
+             plot.title = element_text(hjust = 0.5, vjust = 2))
+
+get_graph_adjacencies <- function(G) {
+  edges <- ends(G, E(G))
+  list(i = edges[, 1], j = edges[, 2])
+}
+
+make_linestring <- function(x1, y1, x2, y2) {
+  st_linestring(matrix(c(x1, y1, x2, y2), ncol = 2, byrow = TRUE))
+}
+
+get_graph_as_line_layer <- function(G, crs) {
+  ij <- get_graph_adjacencies(G)
+  i <- ij$i
+  j <- ij$j
+  edges_as_sf <- mapply(make_linestring, 
+                        V(G)$x[i], V(G)$y[i], V(G)$x[j], V(G)$y[j],
+                        SIMPLIFY = FALSE) |>
+    st_sfc() |>
+    st_as_sf(crs = crs) |>
+    rename(geometry = x)
+  edges_as_sf                      
+}
+
+get_lattice <- function(dist, pts) {
+  pts |>
+    st_buffer(dist) |>
+    st_contains(pts, sparse = FALSE) |>
+    as.matrix() |>
+    graph_from_adjacency_matrix(mode = "undirected", diag = FALSE) |>
+    set_vertex_attr("x", value = pts$x) |>
+    set_vertex_attr("y", value = pts$y)
+}
+
+get_graph_plot <- function(G, main, crs = 2193) {
+  return(ggplot() +
+    geom_sf(data = G |> 
+              get_graph_as_line_layer(crs = 2193)) +
+    geom_sf(data = G |> 
+              vertex_attr() |> as.data.frame() |> st_as_sf(coords = 1:2, crs = crs), 
+            size = 8, fill = "white", colour = "black", shape = 21, lwd = 1) +
+    coord_sf(xlim = c(-270, 270) + 1e6, 
+             ylim = c(-270, 270) + 1e6, expand = FALSE) +
+    ggtitle(main))
+}
+
+grid_pts <- expand_grid(x = -5:5 * 100 + 1e6, y = -5:5 * 100 + 1e6) |>
+  as.data.frame() |>
+  st_as_sf(coords = c("x", "y"), crs = 2193, remove = FALSE)
+
+hex_grid_pts <- grid_pts |>
+  st_bbox() |> 
+  st_make_grid(square = FALSE, what = "centers", cellsize = 107.457) |>
+  st_as_sf() |>
+  rename(geometry = x) |>
+  mutate(geometry = geometry + c(-9, -11)) |>
+  st_set_crs(2193)
+
+xy <- hex_grid_pts |> st_coordinates()
+hex_grid_pts <- hex_grid_pts |>
+  mutate(x = xy[, 1], y = xy[, 2])
+
+graphs <- mapply(
+  get_lattice,
+  list(101, 150, 225, 110),
+  list(grid_pts, grid_pts, grid_pts, hex_grid_pts))
+names(graphs) <- list("Rook's move", "Queen's move", "Knight's move", "Hex")
+
+ggplots <- list()
+for (n in names(graphs)) {
+  ggplots[[n]] <- get_graph_plot(G = graphs[[n]], main = n)
+}
+ggarrange(plotlist = ggplots, ncol = 2, nrow = 2)
+```
+
+The advantage of a square lattice is the ease of deriving vertex elevations assuming a DEM at the desired lattice spacing is available. 
+A hexagonal lattice on the other hand will usually require that elevation values be interpolated from a DEM to the lattice locations.
+
+Both square and hexagonal lattices can lead to geometric artifacts when the allowed moves are used to determine contours of equal travel time or *isochrones* from an origin point. 
+On level terrain the Rook's move produces diamond-shaped isochrones, while the Queen's move produces octagons. 
+In general isochrones will be polygons with as many sides as the number of neighbours of each vertex. 
+While the 16-gons of the Knight's move lattice are likely to be close enough to circular for many purposes, the resulting graph is much larger than the alternatives.
+@Etherington2012 suggests that by combining results from randomly generated planar graphs such geometric artifacts can be removed, but this substantially increases the computational requirements.
+Another computationally intensive approach might draw on methods for
+ estimating geodesics on complex triangular meshes [@Martinez2005].
+Because the geometric effects of the graph structure are masked when we introduce varying movement speeds due to the slope of each edge, we do not consider these more complex approaches necessary in our application.
+Nevertheless, more circular base isochrone shapes are to be preferred, all other things equal.
+
+A further complication when a lattice includes edges of varying length as in the Queen's and Knight's move lattices, is bias in the estimation of slopes leading to lower estimated movement costs for longer edges.
+This is because longer edges (e.g., the knight's moves) may 'jump' across intervening segments of varying slope, smoothing them to a single slope estimate.
+Such smoothing will usually result in estimated speeds along that edge higher than those that would accumulate along the intervening sections of varying slope. 
+This problem is discussed in a slightly different context by Campbell et al. [-@Campbell2019 pp. 96-98].
+
+Any choice of graph structure is necessarily a compromise, and estimates of movement rates are always an approximation [@Goodchild2020]. 
+We consider the differing edge lengths introduced by all the square lattices other than the Rook's case to be problematic, and favour the hexagonal lattice structure over the simple square lattice because a base hexagonal isochrone is preferable to a diamond shape.
+
+Having settled on a hexagonal lattice for the graph structure, at a chosen resolution we set out a regular hexagonal grid of locations across the study area, and assign to each point an elevation by interpolation from a DEM.
+Because the DEM is at finer resolution than the hexagonal lattice, the choice of interpolation method is not a major concern.
+Our approach applies a bilinear interpolation method based on elevation values in the DEM cell the graph vertex falls in and its four orthogonal neighbours.
+Based on the difference in elevation of the vertices at each end of each edge we estimate a slope, and also a traversal times using our estimated hiking function. 
+The important point here is that our graph is _directed_ so that different traversal times $t_{ij}$ and $t_{ji}$ are estimated for moving between locations $i$ to location $j$ depending on the direction of movement.
+
+As noted in the previous section, we estimate two hiking functions, one for moraine and one for rock landcover. 
+When an edge connects locations of two different landcover types the estimated traversal time is the mean of the traversal times for each landcover.
 
@@
 
-[FIGURE showing location information etc., based on these data]
+\[FIGURE showing location information etc., based on these data\]
 
 ### GPS data from an expedition
-
 Devices used, and associated protocols for scientists while on site.
 
-[FIGURE showing GPS traces in context]
+\[FIGURE showing GPS traces in context\]
 
@@
 
-[FIGURE showing filtering steps and impact on slope vs speed scatterplot/boxplots]
+\[FIGURE showing filtering steps and impact on slope vs speed scatterplot/boxplots\]
 
 ## Methods and results
+
 ### Hiking functions {#sec-hiking-function-estimation}
@@
 
-[FIGURE showing hiking function fits about here]
+\[FIGURE showing hiking function fits about here\]
 
@@
 
-[FIGURE comparing moraine and rock hiking functions about here]
+\[FIGURE comparing moraine and rock hiking functions about here\]
 
 ### Landscapes as graphs
-Assignment of nodes as a lattice of equally spaced locations.
-Choice of rectangular/square grid vs.
-hexagonal.
-Square grids are easily generated and also offer some computational advantages if, for example, the grid is aligned with an underlying digital elevation model (DEM).
-However, square grids present the problem of choosing between two different potential local neighbourhoods, based on orthogonal adjacency only, or on orthogonal and diagonal adjacency (respectively referred to as rook and queen’s case adjacency, or as von Neumann and Moore neighbourhoods, in the spatial analysis and cellular automaton literatures).
-Excluding diagonal adjacency yields implausible diamond shaped isochrones on level terrain.
-Including diagonal adjacency can lead to seemingly contradictory results when alternate corners of a square are at different heights.
-
-Given these challenges we adopt a hexagonally arranged triangular lattice.
-This avoids any difficulties with varying distances (i.e.
-network edge lengths) between nodes.
-Note that this is necessarily an approximation, and on extensive flat terrains will yield unrealistic hexagonal isochrones.
-It also presents some computational disadvantages in assigning elevations to graph nodes where interpolation of heights is required due to mismatches between hexagon centres and cells in grid-based DEMs.
-
-[FIGURE showing square and hexagonal lattices and issues]
-
-Alternative approaches might estimate movement based on generating multiple random approximately evenly spaced nodes across a terrain and combining the resulting estimated travel times [see @Etherington2012], or deploying methods for estimating geodesics on complex triangular meshes [@Martinez2005] but both approaches are more computationally demanding than the current application justifies.
-In any case, the latter method does not allow for asymmetry in traversal times as arising due to slopes.
-
 We developed R code [@RCoreTeam2024] to build graphs (i.e.
 networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions.
-Graphs are stored as igraph package [@csardi2006igraph;@Csardi2024] graph objects for further analysis.
+Graphs are stored as igraph package [@csardi2006igraph; @Csardi2024] graph objects for further analysis.
 
@@
 
-`igraph` implements betweenness centrality measures [@Freeman1978;@Brandes2001], which we used to identify vertices and/or edges most likely to be traversed by minimum travel time routes across the terrain.
+`igraph` implements betweenness centrality measures [@Freeman1978; @Brandes2001], which we used to identify vertices and/or edges most likely to be traversed by minimum travel time routes across the terrain.
 
@@
 ### Impact minimizing networks
+
 Tentative proposal for impact minimizing networks based on minimum spanning trees, but noting the issue with respect to directed graphs when these would more correctly be arborescences [@Korte2018a].
@@
 ## Discussion
+
 Blah
@@
 ## Conclusions
+
 Blah

### 2024-12-05 paper/paper.qmd
latest iteration of paper

@@
     toc: true
+    # code-fold: true
     number-sections: true
@@
   message: false
+  cache: true
+  freeze: auto
 ---
+
 ```{css, echo = FALSE}
+/* This is to make things easier to read in the HTML */
 p {color: black;}
@@
 library(sf)
+library(terra)
+library(akima)
+library(stringr)
+library(ggnewscale)
+# library(whitebox)
+library(ggspatial)
 ```
@@
 
-### Hiking functions
+### Hiking functions {#sec-hiking-functions}
 
@@
 
-The functional form of some of these functions is shown in @tbl-hiking-function-equations and graphed in @fig-hiking-functions.
+The functional forms of some hiking functions are shown in @tbl-hiking-function-equations and graphed in @fig-hiking-functions.
 
 | Description | Equation                                 | Examples                        |
-|:--------------------------|-------------------------|:------------------|
+|:-           |:-                                        |:-----                           |
 | Exponential | $ae^{b\left|\theta-c\right|}$            | @Tobler1993, @Marquez-Perez2017 |
@@
 
-: Functional forms of hiking functions {#tbl-hiking-function-equations .striped tbl-colwidths="\[20,20,80\]"}
+: Functional forms of hiking functions {#tbl-hiking-function-equations .striped}
 
@@
 theme_set(theme_minimal())
-theme_update(panel.border = element_rect(linewidth = 0.5, fill = NA))
+theme_update(
+  panel.border = element_rect(linewidth = 0.5, fill = NA))
 
@@
 
-The parameterisations of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the exponential function shown are $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
-These parameters give *Tobler's hiking function* but, as has been noted elsewhere, [see @Herzog2010; @Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds presented by Imhof [-@Imhof1950, pp.
+The parameterisations of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the exponential function shown are such that $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
+These parameters give *Tobler's hiking function* but, as has been noted elsewhere, [see @Herzog2010; @Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds found in Imhof [-@Imhof1950, pp.
 217-220].
-Nevertheless these parameter values are widely applied in the literature.
+Nevertheless, these parameter values are widely applied in the literature.
 
@@
 
-It is commonplace in many applications to also incorporate a penalty on movement contingent on landcover, especially for off-track or off-road movement. 
+It is commonplace in many applications to also incorporate a penalty on movement contingent on land cover, especially for off-track or off-road movement. 
 For example the speed attainable off-track in forested terrain might be only half that attainable in grasslands. 
-Unsurprisingly, there are no widely agreed landcover penalties, but see for example, those compiled by @Herzog2020. 
+Unsurprisingly, there are no widely agreed land cover penalties, but see for example, those compiled by @Herzog2020. 
 A hiking function derived by @Wood2023 includes the local gradient of the terrain (i.e., the maximum slope at each location, not the slope in the direction of movement) as a covariate in the estimated function.
-It is possible that this kind of approach treating landcover as a categorical covariate in the estimation of more complex hiking functions might be more widely applied in future work. 
-In our application, because there are only two kinds of navigable landcover&mdash;moraine (or gravel) and rock&mdash;we chose instead to estimate two hiking functions, one for each landcover, and estimate movement costs conditional on the landcover present at each location.
-This also has the advantage of allowing for differing effects of slope on speed due to landcover, where for example, gravel might allow more rapid movement on the level surfaces, but more rapidly reduce speed on slopes.
+It is possible that this kind of approach including other spatial covariates, such as land cover (which would be a categorical variable in most cases) in the estimation of complex hiking functions might be more widely applied in future work.
+In our application, because there are only two kinds of navigable land cover&mdash;moraine (or gravel) and rock&mdash;we chose instead to estimate two hiking functions, one for each land cover, and estimate movement costs conditional on the land cover at each location.
+This also has the advantage of allowing for differing effects of slope on speed due to land cover, where for example, gravel might allow more rapid movement on the level surfaces, but more rapidly reduce speed on slopes.
 Details of this approach are reported and discussed in @sec-hiking-function-estimation.
@@
 ### Representing the landscape
-Hiking functions are usually asymmetric with respect to slope, with the highest attainable speed at a slight downhill slope. 
+A has been noted, hiking functions are usually asymmetric, with the highest attainable speed at a slight downhill slope. 
 This asymmetry means that unless analysis is focused on assessing movement cost from a single origin or to a single destination, it is necessary to represent the landscape in a way that can accommodate asymmetry.
-We therefore represent the landscape as a directed graph. 
-Graph vertices are locations with associated elevation and landcover. 
-Graph edges are directed links between vertices for which a change in elevation between the start and end vertex can be calculated, and a slope derived, based on the elevation difference and horizontal distance between the vertices.
+We therefore represent the landscape as a directed graph $G(V,E)$ of vertices and directed edges connecting them.
+In this representaiton, graph vertices $v_i$ are locations with associated elevation and land cover. 
+Vertices minimally have spatial coordinates $(x_i,y_i)$, an elevation $z_i$, and a land cover, $C_i$:
 
-The most obvious way to derive such a graph is to assign each cell in a raster digital elevation model (DEM) to a graph vertex, so that graph vertices are arranged in a regular square grid or lattice.
-This approach has been widely adopted **CITATIONS NEEDED HERE** 
-However, this still requires a decision to be made regarding the definition of graph edges, which can be thought of as _allowed moves_ between vertices. 
+$$
+v_i=\left(x_i,y_i,z_i,C_i\right)
+$$ {#eq-vertex}
+
+Graph edges $e_{ij}=\left(v_i,v_j\right)$ are directed connections between vertices for which a change in elevation between the start and end vertex can be calculated, and a slope derived, based on the elevation difference, and the horizontal distance between the vertices.
+Thus the slope $\theta_{ij}$ of edge $e_ij$ is given by
+
+$$
+\theta_{ij}=\frac{z_j-z_i}{\sqrt{(x_i-x_j)^2+(y_i-y_j)^2}}
+$$ {#eq-slope-of-edge}
+
+The obvious way to derive such a graph from spatial data is to assign each cell in a raster digital elevation model (DEM) to a graph vertex, so that graph vertices are arranged in a regular square grid or lattice, and this approach has been widely adopted.
+**CITATIONS NEEDED HERE**
+However, this still requires a decision about how to define graph edges, which can be thought of in terms of _allowed moves_ between vertices. 
 Three possibilities on a square lattice&mdash;Rook's, Queen's, and Knight's moves&mdash;are shown in @fig-lattice-adjacencies. 
-Another option shown in the figure is to lay out a regular hexagonal grid of vertex locations and link each vertex with its six nearest neighbours.
+Another option shown in the figure is to lay out a regular hexagonal grid of vertex locations and link each vertex to its six nearest neighbours.
 
@@
 theme_set(theme_void()) 
-theme_update(plot.margin = unit(c(1, 1, 2, 1), "mm"),
+theme_update(
+  plot.margin = unit(c(1, 1, 2, 1), "mm"),
   plot.title = element_text(hjust = 0.5, vjust = 2))
@@
     as.matrix() |>
-    graph_from_adjacency_matrix(mode = "undirected", diag = FALSE) |>
+    graph_from_adjacency_matrix(mode = "directed", diag = FALSE) |>
     set_vertex_attr("x", value = pts$x) |>
@@
 
-grid_pts <- expand_grid(x = -5:5 * 100 + 1e6, y = -5:5 * 100 + 1e6) |>
+grid_pts <- expand_grid(x = -20:20 * 100 + 1e6, y = -20:20 * 100 + 1e6) |>
   as.data.frame() |>
@@
   rename(geometry = x) |>
-  mutate(geometry = geometry + c(-9, -11)) |>
+  mutate(geometry = geometry + c(-12, -3)) |>
   st_set_crs(2193)
@@
 
-The advantage of a square lattice is the ease of deriving vertex elevations assuming a DEM at the desired lattice spacing is available. 
-A hexagonal lattice on the other hand will usually require that elevation values be interpolated from a DEM to the lattice locations.
+The advantage of a square lattice is that is east to derive vertex elevations assuming a DEM at the desired lattice spacing is available. 
+A hexagonal lattice on the other hand usually requires that elevation values be interpolated from a DEM to the lattice locations.
 
-Both square and hexagonal lattices can lead to geometric artifacts when the allowed moves are used to determine contours of equal travel time or *isochrones* from an origin point. 
-On level terrain the Rook's move produces diamond-shaped isochrones, while the Queen's move produces octagons. 
+Both square and hexagonal lattices lead to geometric artifacts when the allowed moves are used to determine contours of equal travel time or *isochrones* from an origin point on a flat surface with uniform movement speeds.
+In this situation, the Rook's move produces diamond-shaped isochrones, while the Queen's move produces octagons.
 In general isochrones will be polygons with as many sides as the number of neighbours of each vertex. 
-While the 16-gons of the Knight's move lattice are likely to be close enough to circular for many purposes, the resulting graph is much larger than the alternatives.
+While the 16-gons of the Knight's move lattice are likely to be close enough to circular for many purposes, the resulting graph is denser than the alternatives. 
 @Etherington2012 suggests that by combining results from randomly generated planar graphs such geometric artifacts can be removed, but this substantially increases the computational requirements. 
-Another computationally intensive approach might draw on methods for
- estimating geodesics on complex triangular meshes [@Martinez2005].
-Because the geometric effects of the graph structure are masked when we introduce varying movement speeds due to the slope of each edge, we do not consider these more complex approaches necessary in our application.
+Another computationally intensive approach might draw on methods for estimating geodesics on complex triangular meshes [@Martinez2005].
+Because these geometric effects of graph structure are masked when we introduce varying movement speeds due to the slope of each edge, we do not consider these more complex approaches necessary in our application.
 Nevertheless, more circular base isochrone shapes are to be preferred, all other things equal.
@@
 This is because longer edges (e.g., the knight's moves) may 'jump' across intervening segments of varying slope, smoothing them to a single slope estimate.
-Such smoothing will usually result in estimated speeds along that edge higher than those that would accumulate along the intervening sections of varying slope. 
+Such smoothing will usually result in shorter edge traversal times along such edges than those that would accumulate along intervening sections of varying slope. 
 This problem is discussed in a slightly different context by Campbell et al. [-@Campbell2019 pp. 96-98].
@@
 Because the DEM is at finer resolution than the hexagonal lattice, the choice of interpolation method is not a major concern.
-Our approach applies a bilinear interpolation method based on elevation values in the DEM cell the graph vertex falls in and its four orthogonal neighbours.
-Based on the difference in elevation of the vertices at each end of each edge we estimate a slope, and also a traversal times using our estimated hiking function. 
-The important point here is that our graph is _directed_ so that different traversal times $t_{ij}$ and $t_{ji}$ are estimated for moving between locations $i$ to location $j$ depending on the direction of movement.
+Our approach applies bilinear interpolation based on elevation values in the DEM cell the graph vertex falls in and its four orthogonal neighbours.
+Based on the difference in elevation of the vertices at each end of each edge we estimate a slope using @eq-slope-of-edge, and also traversal times using our estimated hiking function. 
+The important point here is that our graph is _directed_ so that different traversal times $t_{ij}$ and $t_{ji}$ are estimated for moving between vertex $v_i$ and $v_j$ depending on the direction of travel.
 
-As noted in the previous section, we estimate two hiking functions, one for moraine and one for rock landcover. 
-When an edge connects locations of two different landcover types the estimated traversal time is the mean of the traversal times for each landcover.
+As noted in the previous section, we estimate two hiking functions, one for moraine and one for rock land cover. 
+When an edge connects locations of two different land cover types the estimated traversal time is the mean of the traversal times for each land cover.
 
@@
 ### Antarctic geospatial data
-@Cox2023, @Cox2023a, @Felden2023
+Geospatial data for Antarctica were obtained from sources referenced in @Cox2023, @Cox2023a, and @Felden2023.
+The study area was defined to be the Skelton and Dry Valleys basins, as defined by the NASA Making Earth System Data Records for Use in Research Environments (MEaSUREs) project [@Mouginot2017] and shown in @fig-locator-map.
+The Skelton basin was included because while the expedition GPS data was ostensibly collected in the McMurdo Dry Valleys, it actually extends into that basin as shown in @fig-study-area-map.
+Elevation data from the Reference Elevation Model of Antarctica project @Howat2022, and geology from GeoMAP @Cox2023 are shown in @fig-dem-and-geology.
+The five largest areas of contiguous non-ice surface geology across the study area shown in @fig-five-subregions were chosen to be the specific sites for more detailed exploration using the methods set out in this paper.
+These range in size from around 320 to 2600 square kilometres.
 
-\[FIGURE showing location information etc., based on these data\]
+```{r}
+#| output: false
+data_folder <- "/Users/david/Documents/work/mwlr-tpm-antarctica/antarctica/_data"
+dem_folder  <- "/Users/david/Documents/work/mwlr-tpm-antarctica/antarctica/_data/dem"
+
+coastline <- st_read(str_glue("{data_folder}/coastline.gpkg"))
+basins <- st_read(str_glue("{dem_folder}/antarctic-basins.gpkg"))
+study_basins <- st_read(str_glue("{dem_folder}/dry-valleys-and-skelton.gpkg"))
+study_area <- st_read(str_glue("{dem_folder}/study-area-extent.gpkg"))
+gps_data <- st_read(str_glue("{data_folder}/cleaned-gps-data/all-gps-traces.gpkg"))
+tiles <- st_read(str_glue("{dem_folder}/REMA-index-included-tiles.gpkg"))
+shade <- rast(str_glue("{dem_folder}/32m/dry-valleys-combined-32m-browse.tif"))
+shade8 <- shade |>
+  terra::aggregate(8) |>
+  as.data.frame(xy = TRUE) |>
+  rename(shade = `15_35_32m_v2.0_browse`)
+geology <- st_read(str_glue("{data_folder}/ata-scar-geomap-geology-v2022-08-clipped.gpkg"))
+contiguous_geologies <- st_read(str_glue("{data_folder}/dry-valleys/common-inputs/contiguous-geologies.gpkg"))
+```
+
+::: {#fig-study-area layout="[[50,-1,50], [50,-1,50]]"}
+
+```{r}
+#| label: fig-locator-map
+#| fig-cap: "Study area location in Antarctica"
+theme_set(theme_minimal())
+theme_update(
+  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
+  panel.grid = element_line(colour = "black", linewidth = 0.05),
+  panel.ontop = TRUE
+)
+bb1 <- st_bbox(study_area)
+ggplot() +
+  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
+  geom_sf(data = study_area, fill = "#666666", colour = "white") +
+  annotate("rect", xmin = bb1[1], ymin = bb1[2], 
+                   xmax = bb1[3], ymax = bb1[4], fill = "#00000000", colour = "red", lwd = 0.25)
+```
+
+```{r}
+#| label: fig-study-area-map
+#| fig-cap: "Skelton and Dry Valleys basins"
+theme_set(theme_void())
+theme_update(
+  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5)
+)
+ggplot() +
+  geom_sf(data = coastline, fill = "#eff3ff", colour = "#00000000") +
+  geom_sf(data = study_basins, fill = "gray", colour = "white") +
+  geom_sf(data = coastline, fill = "#00000000", colour = "#3182bd", lwd = 0.35) +
+  geom_sf(data = gps_data, colour = "red", size = 0.025) +
+  geom_sf_text(data = study_basins, aes(label = NAME)) +
+  coord_sf(xlim = bb1[c(1, 3)], ylim = bb1[c(2, 4)]) +
+  annotation_scale(height = unit(0.1, "cm"), location = "tr") +
+  annotation_north_arrow(style = north_arrow_minimal, which_north = "true", location = "br")
+```
+
+```{r}
+#| label: fig-dem-and-geology
+#| fig-cap: "Study area elevation (hillshade) and surface geology"
+theme_set(theme_void())
+theme_update(
+  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
+  legend.position = c(0.02, 0.02),
+  legend.justification = c(0, 0)
+)
+top5 <- contiguous_geologies |> 
+  arrange(desc(area)) |> 
+  slice(1:5)
+bb2 <- st_bbox(top5)
+ggplot() +
+  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd", lwd = 0.35) +
+  geom_sf(data = study_area, fill = "#00000000", colour = "darkblue", lwd = 0.35) +
+  geom_sf(data = geology, aes(fill = POLYGTYPE), linewidth = 0) +
+  scale_fill_manual(values = c("dodgerblue", "#fec44f", "#993404"), name = "Geology") + 
+  new_scale("fill") +
+  geom_raster(data = shade8, aes(x = x, y = y, fill = shade), alpha = 0.5) +
+  scale_fill_distiller(palette = "Blues", guide = "none") +
+  annotate("rect", xmin = bb2[1], ymin = bb2[2], 
+                   xmax = bb2[3], ymax = bb2[4], fill = "#00000000", 
+                   colour = "red", lwd = 0.25, linetype = "dashed") +
+  coord_sf(xlim = bb1[c(1, 3)], ylim = bb1[c(2, 4)])
+```
+
+```{r}
+#| label: fig-five-subregions
+#| fig-cap: "Five sub-regions of contiguous surface geology"
+theme_set(theme_void())
+theme_update(
+  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5)
+)
+ggplot() +
+  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
+  geom_sf(data = contiguous_geologies, fill = "#eeeeee", colour = "darkgrey") +
+  geom_sf(data = top5, fill = "#fcbba1", colour = "black", linewidth = 0.25) +
+  guides(fill = "none") +
+  annotation_scale(height = unit(0.1, "cm")) +
+  coord_sf(xlim = bb2[c(1, 3)], ylim = bb2[c(2, 4)])
+```
+
+The study area.
+:::
 
 ### GPS data from an expedition
-Devices used, and associated protocols for scientists while on site.
+**FRASER**: Timeline, devices used, and associated protocols for scientists while on site.
 
-\[FIGURE showing GPS traces in context\]
+GPS data were processed (see @fig-gps-processing) to make them better suited for use in the estimation of hiking functions.
 
-Initial filtering to remove implausibly high-speed movement fixes – these largely related to helicopter movements and arrival/departure from the valley.
+The first processing step was to confirm the plausibility of the data, particularly the device-generated speed distance between fixes, and elevations associated with fixes.
+The challenges of post-processing GPS data are well documented and relate to issues with GPS drift which can lead to estimated non-zero movement speeds as a result of noise in the signal.
+The raw GPS data included distance since last fix, speed, and elevation estimates and it was determined in all cases that the device generated results for these measurements were likely to be more reliable than post-processing the raw latitude-longitude fixes to calculate the values.
 
-Additional cleanup of GPS data required: filtering to remove many GPS fixes with associated low movement rates associated with time spent at base camps, rest stops, or experimental sites.
+The second processing step was to remove fixes associated with faster movement on other modes of transport than walking.
+@Wood2023 cite a number of previous works that base detection of trip segments based on recorded speeds.
+This method was trivially applicable to our data to a limited degree as scientists arrive at the expedition base camp and occasionally also travel on helicopters on trips to more remote experimental sites.
 
-Crude filters: short distances between fixes, large turn angles between fixes, and seemingly anomalous high speeds recorded at 0 slope angles (yet to be explained but required to obtain reasonable results).
+The third, more challenging processing step was to deal with sequences of fixes associated with non-purposeful movement when scientists were in or around base camp, at rest stops, or at experimental sites.
+Crude filters removed fixes with low recorded distances between fixes (less than 2.5 metres), high turn angles at the fix location (greater than 150&deg;), and fixes recorded on ice-covered terrain, but this didi not clean the data sufficiently for further analysis.
+An additional filtering step was to count fixes (across all scientists) in square grid cells and remove all fixes in grid cells with more than 50 fixes.
 
-Density-based filtering: hexbinning of fixes to identify non-purposive movement areas.
+This left one persistent concern: an over-representation of consecutive fixes recorded at exactly the same elevation, resulting in many fixes with estimated slopes of exactly 0, and leading to a clearly evident dip in estimated movement speeds at 0 slope (@fig-raw-gps-data).
+It is likely that these fixes are associated with GPS device drift, so a it was decided to remove all fixes where estimated slope was exactly 0.
+@fig-filtered-gps-data shows the improvement in even a crudely estimated hiking function derived from local scatterplot (LOESS) smoothing.
+Note that such functions are likely overfitted and not used further in our analysis where we favour more easily parameterised functions such as those discussed in @sec-hiking-functions.
 
-\[FIGURE showing filtering steps and impact on slope vs speed scatterplot/boxplots\]
+```{r}
+#| output: false
+# gps data and its bbox
+gps_data <- st_read(str_glue("{data_folder}/cleaned-gps-data/all-gps-traces.gpkg"))
+bb <- gps_data |> st_bbox() # this is to allow plotting restricted to GPS extents
+
+# geologies data
+geologies <- 
+  st_read(str_glue("{data_folder}/ata-scar-geomap-geology-v2022-08-clipped.gpkg")) |> 
+  dplyr::select(POLYGTYPE) |>
+  st_filter(bb |> st_as_sfc()) |> 
+  mutate(cover = factor(POLYGTYPE))
+
+# join geologies to the gps data
+gps_geol <- gps_data |>
+  st_join(geologies) |>
+  dplyr::select(-POLYGTYPE) |> 
+  # we do not further consider 'ice' terrain, or
+  # observations at negative elevations
+  filter(cover != "ice", height_m > 0) |>
+  mutate(slope_h_round = round(slope_h * 10) / 10) |>
+  drop_na()
+```
+
+::: {#fig-gps-data layout="[[50,-1,50]]"}
+
+```{r}
+#| label: fig-raw-gps-data
+#| fig-cap: "Boxplots by slope of speed, with smoothed estimated hiking function showing a 'dip' due to over-representation of 0 slope fixes"
+ggplot(gps_geol |> filter(slope_h >= -1, slope_h <= 1.0)) +
+  geom_boxplot(aes(x = round(slope_h, 1), y = speed_km_h, group = round(slope_h, 1)), 
+               outlier.size = 0.2, colour = "grey", linewidth = 0.5) +
+  stat_smooth(aes(x = slope_h, y = speed_km_h)) +
+  xlab("Slope, rise over run") + ylab("Speed, km/h") +
+  theme_minimal()
+```
+
+```{r}
+#| label: fig-filtered-gps-data
+#| fig-cap: "After filtering the estimated hiking function no longer has a dip."
+hexes <- gps_geol |> 
+  st_make_grid(cellsize = 100, square = FALSE, what = "polygons") |>
+  st_as_sf(crs = st_crs(gps_geol)) |>
+  rename(geom = x) |>
+  st_join(gps_geol |> mutate(id = row_number()), left = FALSE) |>
+  group_by(geom) |>
+  summarise(n = n(), n_persons = n_distinct(name)) |>
+  ungroup()
+
+gps_geol_purposive <- gps_geol |> 
+  # see above
+  filter(slope_h != 0) |>
+  # these two remove 'dithering'
+  filter(turn_angle < 150) |>
+  filter(distance_m > 2.5) |>
+  # remove fixes in densely trafficked areas
+  st_join(hexes) |>
+    filter(n <= 50)
+
+ggplot(gps_geol_purposive) +
+  geom_boxplot(aes(x = round(slope_h, 1), y = speed_km_h, group = round(slope_h, 1)), 
+               outlier.size = 0.2, colour = "grey", linewidth = 0.5) +
+  stat_smooth(aes(x = slope_h, y = speed_km_h)) +
+  xlab("Slope, rise over run") + ylab("Speed, km/h") +
+  theme_minimal()
+```
+
+GPS data and a crudely estimated hiking functions before and after filtering of the data.
+:::
 

### 2024-12-05 paper/paper_.qmd
latest iteration of paper

--- 2024-12-05 paper/paper.qmd
@@
 
-### Hiking functions {#sec-hiking-functions}
+### Hiking functions
 
@@
 The study area was defined to be the Skelton and Dry Valleys basins, as defined by the NASA Making Earth System Data Records for Use in Research Environments (MEaSUREs) project [@Mouginot2017] and shown in @fig-locator-map.
-The Skelton basin was included because while the expedition GPS data was ostensibly collected in the McMurdo Dry Valleys, it actually extends into that basin as shown in @fig-study-area-map.
-Elevation data from the Reference Elevation Model of Antarctica project @Howat2022, and geology from GeoMAP @Cox2023 are shown in @fig-dem-and-geology.
-The five largest areas of contiguous non-ice surface geology across the study area shown in @fig-five-subregions were chosen to be the specific sites for more detailed exploration using the methods set out in this paper.
-These range in size from around 320 to 2600 square kilometres.
+The Skelton basin was included because, while the expedition GPS data was ostensibly collected in the McMurdo Dry Valleys, it actually extends into that basin as shown in @fig-study-area-map.
+@fig-dem-and-geology shows the study area
 
@@
 
-::: {#fig-study-area layout="[[50,-1,50], [50,-1,50]]"}
+::: {#fig-study-area layout-ncol=2 layout-nrow=2}
 
@@
 #| label: fig-locator-map
-#| fig-cap: "Study area location in Antarctica"
+#| fig-cap: "Study area location"
 theme_set(theme_minimal())
@@
   panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
+  panel.background = element_rect(fill = "#6baed6"),
   panel.grid = element_line(colour = "black", linewidth = 0.05),
-  panel.ontop = TRUE
+  # panel.ontop = TRUE
 )
-bb1 <- st_bbox(study_area)
+bb <- st_bbox(study_area)
 ggplot() +
@@
   geom_sf(data = study_area, fill = "#666666", colour = "white") +
-  annotate("rect", xmin = bb1[1], ymin = bb1[2], 
-                   xmax = bb1[3], ymax = bb1[4], fill = "#00000000", colour = "red", lwd = 0.25)
+  annotate("rect", xmin = bb[1], ymin = bb[2], xmax = bb[3], ymax = bb[4], fill = "#00000000", colour = "red")
 ```
@@
 ggplot() +
-  geom_sf(data = coastline, fill = "#eff3ff", colour = "#00000000") +
+  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
   geom_sf(data = study_basins, fill = "gray", colour = "white") +
-  geom_sf(data = coastline, fill = "#00000000", colour = "#3182bd", lwd = 0.35) +
   geom_sf(data = gps_data, colour = "red", size = 0.025) +
   geom_sf_text(data = study_basins, aes(label = NAME)) +
-  coord_sf(xlim = bb1[c(1, 3)], ylim = bb1[c(2, 4)]) +
+  coord_sf(xlim = bb[c(1, 3)], ylim = bb[c(2, 4)]) +
   annotation_scale(height = unit(0.1, "cm"), location = "tr") +
-  annotation_north_arrow(style = north_arrow_minimal, which_north = "true", location = "br")
+  annotation_north_arrow(style = north_arrow_minimal, which_north = "true", where = "br")
 ```
@@
 #| label: fig-dem-and-geology
-#| fig-cap: "Study area elevation (hillshade) and surface geology"
-theme_set(theme_void())
-theme_update(
-  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
-  legend.position = c(0.02, 0.02),
-  legend.justification = c(0, 0)
-)
-top5 <- contiguous_geologies |> 
-  arrange(desc(area)) |> 
-  slice(1:5)
-bb2 <- st_bbox(top5)
-ggplot() +
-  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd", lwd = 0.35) +
-  geom_sf(data = study_area, fill = "#00000000", colour = "darkblue", lwd = 0.35) +
-  geom_sf(data = geology, aes(fill = POLYGTYPE), linewidth = 0) +
-  scale_fill_manual(values = c("dodgerblue", "#fec44f", "#993404"), name = "Geology") + 
-  new_scale("fill") +
-  geom_raster(data = shade8, aes(x = x, y = y, fill = shade), alpha = 0.5) +
-  scale_fill_distiller(palette = "Blues", guide = "none") +
-  annotate("rect", xmin = bb2[1], ymin = bb2[2], 
-                   xmax = bb2[3], ymax = bb2[4], fill = "#00000000", 
-                   colour = "red", lwd = 0.25, linetype = "dashed") +
-  coord_sf(xlim = bb1[c(1, 3)], ylim = bb1[c(2, 4)])
-```
-
-```{r}
-#| label: fig-five-subregions
-#| fig-cap: "Five sub-regions of contiguous surface geology"
+#| fig-cap: "Study area geology and hillshade"
 theme_set(theme_void())
@@
 ggplot() +
+  geom_raster(data = shade8, aes(x = x, y = y, fill = shade), alpha = 0.5) +
+  scale_fill_distiller(palette = "Blues", guide = "none") +
   geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
-  geom_sf(data = contiguous_geologies, fill = "#eeeeee", colour = "darkgrey") +
-  geom_sf(data = top5, fill = "#fcbba1", colour = "black", linewidth = 0.25) +
+  geom_sf(data = study_area, fill = "#00000000", colour = "red", lwd = 0.35) +
+  new_scale("fill") +
+  geom_sf(data = geology, aes(fill = POLYGTYPE), linewidth = 0) +
+  scale_fill_manual(values = c("dodgerblue", "grey", "#666666"), name = "Geology") + 
+  coord_sf(xlim = bb[c(1, 3)], ylim = bb[c(2, 4)])
+```
+
+```{r}
+#| label: fig-five-subregions
+#| fig-cap: "Five sub-areas of contiguous surface geology"
+theme_set(theme_void())
+theme_update(
+  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5)
+)
+top5 <- contiguous_geologies |> arrange(desc(area)) |> slice(1:5)
+bb <- st_bbox(top5)
+ggplot() +
+  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
+  geom_sf(data = contiguous_geologies, fill = "#00000000", colour = "darkgrey") +
+  geom_sf(data = top5, fill = "#00000000", colour = "black", linewidth = 0.25) +
   guides(fill = "none") +
   annotation_scale(height = unit(0.1, "cm")) +
-  coord_sf(xlim = bb2[c(1, 3)], ylim = bb2[c(2, 4)])
+  coord_sf(xlim = bb[c(1, 3)], ylim = bb[c(2, 4)])
 ```
@@
 ### GPS data from an expedition
-**FRASER**: Timeline, devices used, and associated protocols for scientists while on site.
+Devices used, and associated protocols for scientists while on site.
 
-GPS data were processed (see @fig-gps-processing) to make them better suited for use in the estimation of hiking functions.
+\[FIGURE showing GPS traces in context\]
 
-The first processing step was to confirm the plausibility of the data, particularly the device-generated speed distance between fixes, and elevations associated with fixes.
-The challenges of post-processing GPS data are well documented and relate to issues with GPS drift which can lead to estimated non-zero movement speeds as a result of noise in the signal.
-The raw GPS data included distance since last fix, speed, and elevation estimates and it was determined in all cases that the device generated results for these measurements were likely to be more reliable than post-processing the raw latitude-longitude fixes to calculate the values.
+Initial filtering to remove implausibly high-speed movement fixes – these largely related to helicopter movements and arrival/departure from the valley.
 
-The second processing step was to remove fixes associated with faster movement on other modes of transport than walking.
-@Wood2023 cite a number of previous works that base detection of trip segments based on recorded speeds.
-This method was trivially applicable to our data to a limited degree as scientists arrive at the expedition base camp and occasionally also travel on helicopters on trips to more remote experimental sites.
+Additional cleanup of GPS data required: filtering to remove many GPS fixes with associated low movement rates associated with time spent at base camps, rest stops, or experimental sites.
 
-The third, more challenging processing step was to deal with sequences of fixes associated with non-purposeful movement when scientists were in or around base camp, at rest stops, or at experimental sites.
-Crude filters removed fixes with low recorded distances between fixes (less than 2.5 metres), high turn angles at the fix location (greater than 150&deg;), and fixes recorded on ice-covered terrain, but this didi not clean the data sufficiently for further analysis.
-An additional filtering step was to count fixes (across all scientists) in square grid cells and remove all fixes in grid cells with more than 50 fixes.
+Crude filters: short distances between fixes, large turn angles between fixes, and seemingly anomalous high speeds recorded at 0 slope angles (yet to be explained but required to obtain reasonable results).
 
-This left one persistent concern: an over-representation of consecutive fixes recorded at exactly the same elevation, resulting in many fixes with estimated slopes of exactly 0, and leading to a clearly evident dip in estimated movement speeds at 0 slope (@fig-raw-gps-data).
-It is likely that these fixes are associated with GPS device drift, so a it was decided to remove all fixes where estimated slope was exactly 0.
-@fig-filtered-gps-data shows the improvement in even a crudely estimated hiking function derived from local scatterplot (LOESS) smoothing.
-Note that such functions are likely overfitted and not used further in our analysis where we favour more easily parameterised functions such as those discussed in @sec-hiking-functions.
+Density-based filtering: hexbinning of fixes to identify non-purposive movement areas.
 
-```{r}
-#| output: false
-# gps data and its bbox
-gps_data <- st_read(str_glue("{data_folder}/cleaned-gps-data/all-gps-traces.gpkg"))
-bb <- gps_data |> st_bbox() # this is to allow plotting restricted to GPS extents
-
-# geologies data
-geologies <- 
-  st_read(str_glue("{data_folder}/ata-scar-geomap-geology-v2022-08-clipped.gpkg")) |> 
-  dplyr::select(POLYGTYPE) |>
-  st_filter(bb |> st_as_sfc()) |> 
-  mutate(cover = factor(POLYGTYPE))
-
-# join geologies to the gps data
-gps_geol <- gps_data |>
-  st_join(geologies) |>
-  dplyr::select(-POLYGTYPE) |> 
-  # we do not further consider 'ice' terrain, or
-  # observations at negative elevations
-  filter(cover != "ice", height_m > 0) |>
-  mutate(slope_h_round = round(slope_h * 10) / 10) |>
-  drop_na()
-```
-
-::: {#fig-gps-data layout="[[50,-1,50]]"}
-
-```{r}
-#| label: fig-raw-gps-data
-#| fig-cap: "Boxplots by slope of speed, with smoothed estimated hiking function showing a 'dip' due to over-representation of 0 slope fixes"
-ggplot(gps_geol |> filter(slope_h >= -1, slope_h <= 1.0)) +
-  geom_boxplot(aes(x = round(slope_h, 1), y = speed_km_h, group = round(slope_h, 1)), 
-               outlier.size = 0.2, colour = "grey", linewidth = 0.5) +
-  stat_smooth(aes(x = slope_h, y = speed_km_h)) +
-  xlab("Slope, rise over run") + ylab("Speed, km/h") +
-  theme_minimal()
-```
-
-```{r}
-#| label: fig-filtered-gps-data
-#| fig-cap: "After filtering the estimated hiking function no longer has a dip."
-hexes <- gps_geol |> 
-  st_make_grid(cellsize = 100, square = FALSE, what = "polygons") |>
-  st_as_sf(crs = st_crs(gps_geol)) |>
-  rename(geom = x) |>
-  st_join(gps_geol |> mutate(id = row_number()), left = FALSE) |>
-  group_by(geom) |>
-  summarise(n = n(), n_persons = n_distinct(name)) |>
-  ungroup()
-
-gps_geol_purposive <- gps_geol |> 
-  # see above
-  filter(slope_h != 0) |>
-  # these two remove 'dithering'
-  filter(turn_angle < 150) |>
-  filter(distance_m > 2.5) |>
-  # remove fixes in densely trafficked areas
-  st_join(hexes) |>
-    filter(n <= 50)
-
-ggplot(gps_geol_purposive) +
-  geom_boxplot(aes(x = round(slope_h, 1), y = speed_km_h, group = round(slope_h, 1)), 
-               outlier.size = 0.2, colour = "grey", linewidth = 0.5) +
-  stat_smooth(aes(x = slope_h, y = speed_km_h)) +
-  xlab("Slope, rise over run") + ylab("Speed, km/h") +
-  theme_minimal()
-```
-
-GPS data and a crudely estimated hiking functions before and after filtering of the data.
-:::
+\[FIGURE showing filtering steps and impact on slope vs speed scatterplot/boxplots\]
 

### 2024-12-10 paper/paper.qmd
fixed minor typo
latest edits to paper draft

@@
 library(ggspatial)
+library(minpack.lm)
 ```
@@
 
-GPS data were processed (see @fig-gps-processing) to make them better suited for use in the estimation of hiking functions.
+GPS data were processed to make them better suited for use in the estimation of hiking functions.
 
@@
   st_filter(bb |> st_as_sfc()) |> 
-  mutate(cover = factor(POLYGTYPE))
+  mutate(cover = factor(str_to_title(POLYGTYPE)))
 
@@
 
-GPS data and a crudely estimated hiking functions before and after filtering of the data.
+GPS data and crudely estimated hiking functions before and after filtering the to remove fixes associated iwht non-purposive movement.
 :::
@@
 ## Methods and results
+### Hiking functions {#sec-hiking-function-estimation}
+We fit three alternative functional forms to the cleaned GPS data: exponential [@Tobler1993], Gaussian [following @Irmischer2018], and Lorentz [following @Campbell2019] using the Levenburg-Marquardt algorithm [@More1978] as provided by the `nlsLM` function in the `minpack.lm` R package [@Elzhov2022].
+The raw data and fitted curves are shown in @fig-comparison-of-hiking-functions.
 
-### Hiking functions {#sec-hiking-function-estimation}
-Fit three alternative functional forms: exponential Tobler (1993, pp 1–4, based on data from Imhof 1950), Gaussian Irmischer and Clarke (2018), and Student’s t distribution, the last of these as a heavier-tailed alternative to Gaussian.
-Other approaches rely on rule-of-thumb hiking time calculations such as...
+```{r}
+get_gaussian_hiking_function <- function(df) {
+  nlsLM(speed_km_h ~ a * dnorm(slope_h, m, s), data = df,
+        start = c(a = 5, m = 0, s = 0.5))
+}
+get_tobler_hiking_function <- function(df) {
+  nlsLM(speed_km_h ~ a * exp(-b * abs(slope_h + c)), data = df,
+        start = c(a = 5, b = 3, c = 0.05))  
+}
+get_lorentz_hiking_function <- function(df) {
+  nlsLM(speed_km_h ~ a /(pi * b * (1 + ((slope_h - c) / b) ^ 2)), 
+        data = df, start = c(a = 5, b = 1, c = -0.05))
+}
 
-The Gaussian form offered the best overall fits to our data. 
-Although Tobler’s exponential form is much used it is somewhat implausible that hiking speeds would peak sharply at a very specific slope angle, and so we adopt the Gaussian form.
+# makes a prediction data frame with x, y values
+# slopes is a DF with one column called slope_h
+get_model_prediction_df <- function(m, slopes) {
+  data.frame(x = slopes, y = predict(m, data.frame(slope_h = slopes$slope_h)))
+}
 
-\[FIGURE showing hiking function fits about here\]
+slopes <- data.frame(slope_h = -150:150 / 100)
+
+hiking_functions <- list(
+  Exponential = get_tobler_hiking_function,
+  Gaussian = get_gaussian_hiking_function,
+  Lorentz = get_lorentz_hiking_function
+)
+covers <- c("Moraine", "Rock")
+
+models <- list()
+predictions <- list()
+inputs <- list()
+i <- 1
+for (name in names(hiking_functions)) {
+  for (cover_type in covers) {
+    df <- gps_geol_purposive |> 
+      filter(cover == cover_type) 
+    m <- hiking_functions[[name]](df)
+    models[[cover_type]][[name]] <- m
+    predictions[[i]] <- get_model_prediction_df(m, slopes) |>
+      mutate(cover = cover_type, Model = name)
+    inputs[[i]] <- df |>
+      mutate(cover = cover_type, Model = name)
+    i <- i + 1
+  }
+}
+model_predictions <- bind_rows(predictions)
+input_data <- bind_rows(inputs)
+```
+
+```{r}
+#| label: fig-comparison-of-hiking-functions
+#| fig-cap: Three possible hiking functions applied to GPS data split by land cover.
+ggplot() +
+  geom_boxplot(data = gps_geol_purposive |> filter(cover != "Ice"),
+               aes(x = slope_h_round, y = speed_km_h, group = slope_h_round),
+               outlier.size = 0.35, colour = "grey", linewidth = 0.5) +
+  # geom_point(data = input_data, aes(x = slope_h, y = speed_km_h)) + 
+  geom_line(data = model_predictions, aes(x = slope_h, y = y, colour = Model)) +
+  xlab("Slope, rise over run") + ylab("Speed, km/h") +
+  facet_wrap( ~ cover, ncol = 3) +
+  theme_minimal()
+```
+
+The Lorentz function offers a marginal improvement in the model fit in comparison with the Gaussian function, while both are clearly better than the exponential form. 
+However, the improvement offered by the Lorentz function over the Gaussian is marginal: residual standard error 1.489 vs 1.491 on Moraine, and 1.487 vs 1.488 on Rock, and inspection of the curves shows that estimated hiking speeds for the Gaussian functions are much closer to a plausible zero on very steep slopes.
+We therefore chose to adopt Gaussian hiking functions for the remainder of the present work.
 
 In previous work researchers have applied a ground cover penalty cost to a base hiking function to estimate traversal times.
-In our case there are effectively only two terrain types, moraines (gravel) and bare rock. 
-This makes the option of fitting different functions to each attractive and reasonable, and the results of fitting distinct functions to each support this approach. 
-The peak speed on bare rock is attained on steeper downhill slopes than on gravel, perhaps indicative of the greater care required on downhill gravel slopes. 
-Meanwhile the highest speeds on level terrain are attained on gravel. 
-Our fitted functions are shown below.
+We instead, as shown, estimate different hiking functions for the two ground cover types present.
+The peak speed on rock is attained on steeper downhill slopes than on moraines, perhaps indicative of the greater care required on downhill gravel slopes. 
+Meanwhile the highest speeds on level terrain are attained on moraines.
 
-\[FIGURE comparing moraine and rock hiking functions about here\]
+```{r}
+# make a all data function
+df <- gps_geol_purposive 
+m3 <- get_gaussian_hiking_function(df)
+models[["All"]][["Gaussian"]] <- m
+model_predictions <- model_predictions |>
+  bind_rows(get_model_prediction_df(m, slopes) |> 
+              mutate(cover = "All", model = "Gaussian"))
+input_data <- input_data |>
+  bind_rows(df |> mutate(cover = "All", model = "Gaussian"))
+
+```
+
+```{r}
+#| output: false
+
+# see https://rmazing.wordpress.com/2013/08/14/predictnls-part-1-monte-carlo-simulation-confidence-intervals-for-nls-models/
+predictNLS <- function(object, newdata, level = 0.95, nsim = 10000, ...) {
+  require(MASS, quietly = TRUE)
+  ## get right-hand side of formula
+  RHS <- as.list(object$call$formula)[[3]]
+  EXPR <- as.expression(RHS)
+  ## all variables in model
+  VARS <- all.vars(EXPR)
+  ## coefficients
+  COEF <- coef(object)
+  ## extract predictor variable    
+  predNAME <- setdiff(VARS, names(COEF))  
+  ## take fitted values, if 'newdata' is missing
+  if (missing(newdata)) {
+    newdata <- eval(object$data)[predNAME]
+    colnames(newdata) <- predNAME
+  }
+  ## check that 'newdata' has same name as predVAR
+  if (names(newdata)[1] != predNAME) stop("newdata should have name '", predNAME, "'!")
+  ## get parameter coefficients
+  COEF <- coef(object)
+  ## get variance-covariance matrix
+  VCOV <- vcov(object)
+  ## augment variance-covariance matrix for 'mvrnorm' 
+  ## by adding a column/row for 'error in x'
+  NCOL <- ncol(VCOV)
+  ADD1 <- c(rep(0, NCOL))
+  ADD1 <- matrix(ADD1, ncol = 1)
+  colnames(ADD1) <- predNAME
+  VCOV <- cbind(VCOV, ADD1)
+  ADD2 <- c(rep(0, NCOL + 1))
+  ADD2 <- matrix(ADD2, nrow = 1)
+  rownames(ADD2) <- predNAME
+  VCOV <- rbind(VCOV, ADD2) 
+  ## iterate over all entries in 'newdata' as in usual 'predict.' functions
+  NR <- nrow(newdata)
+  respVEC <- numeric(NR)
+  seVEC <- numeric(NR)
+  varPLACE <- ncol(VCOV)   
+  ## define counter function
+  counter <- function (i) {
+    if (i%%10 == 0) 
+      cat(i)
+    else cat(".")
+    if (i%%50 == 0) 
+      cat("\n")
+    flush.console()
+  }
+  outMAT <- NULL 
+  for (i in 1:NR) {
+    counter(i)
+    ## get predictor values and optional errors
+    predVAL <- newdata[i, 1]
+    if (ncol(newdata) == 2) predERROR <- newdata[i, 2] else predERROR <- 0
+    names(predVAL) <- predNAME  
+    names(predERROR) <- predNAME  
+    ## create mean vector for 'mvrnorm'
+    MU <- c(COEF, predVAL)
+    ## create variance-covariance matrix for 'mvrnorm'
+    ## by putting error^2 in lower-right position of VCOV
+    newVCOV <- VCOV
+    newVCOV[varPLACE, varPLACE] <- predERROR^2
+    ## create MC simulation matrix
+    simMAT <- mvrnorm(n = nsim, mu = MU, Sigma = newVCOV, empirical = TRUE)
+    ## evaluate expression on rows of simMAT
+    EVAL <- try(eval(EXPR, envir = as.data.frame(simMAT)), silent = TRUE)
+    if (inherits(EVAL, "try-error")) stop("There was an error evaluating the simulations!")
+    ## collect statistics
+    PRED <- data.frame(predVAL)
+    colnames(PRED) <- predNAME   
+    FITTED <- predict(object, newdata = data.frame(PRED))
+    MEAN.sim <- mean(EVAL, na.rm = TRUE)
+    SD.sim <- sd(EVAL, na.rm = TRUE)
+    MEDIAN.sim <- median(EVAL, na.rm = TRUE)
+    MAD.sim <- mad(EVAL, na.rm = TRUE)
+    QUANT <- quantile(EVAL, c((1 - level)/2, level + (1 - level)/2))
+    RES <- c(FITTED, MEAN.sim, SD.sim, MEDIAN.sim, MAD.sim, QUANT[1], QUANT[2])
+    outMAT <- rbind(outMAT, RES)
+  }
+  colnames(outMAT) <- c("fit", "mean", "sd", "median", "mad", names(QUANT[1]), names(QUANT[2]))
+  rownames(outMAT) <- NULL
+  cat("\n")
+  return(outMAT)  
+}
+
+conf <- 0.95
+pred1 <- predictNLS(models[["Moraine"]][["Gaussian"]], newdata = slopes, level = conf) |>
+  as.data.frame() |>
+  mutate(x = slopes$slope_h, cover = "Moraine")
+pred2 <- predictNLS(models[["Rock"]][["Gaussian"]], newdata = slopes, level = conf) |>
+  as.data.frame() |>
+  mutate(x = slopes$slope_h, cover = "Rock")
+pred3 <- predictNLS(m3, newdata = slopes, level = conf) |>
+  as.data.frame() |>
+  mutate(x = slopes$slope_h, cover = "All")
+predictions <- bind_rows(pred1, pred2, pred3)
+```
+
+Overplotting of the hiking functions including an additional model fitted to all the data, confirms that the fitted functions are sufficiently different to retain separate models for each ground cover (see @fig-models-by-cover-compared).
+Plotting both functions in the same graph makes clearer the difference in maximum speed and slope at maximum speed associated with each ground cover.
+
+```{r}
+#| label: fig-models-by-cover-compared
+#| fig-cap: The hiking functions for All, Moraine and Rock ground covers compared, including 95% confidence intervals derived by Monte-Carlo simulation.
+ggplot() +
+  geom_ribbon(data = predictions |> filter(cover != "All"), 
+              aes(x = x, ymin = `2.5%`, ymax = `97.5%`, group = cover, fill = cover), 
+              alpha = 0.35, linewidth = 0) + 
+  scale_fill_brewer(palette = "Set1", name = "Cover") +
+  geom_ribbon(data = predictions |> filter(cover == "All"), 
+              aes(x = x, ymin = `2.5%`, ymax = `97.5%`), 
+              colour = "black", lty = "dashed", fill = "#00000000", linewidth = 0.35) +
+  geom_line(data = predictions |> filter(cover == "All"), aes(x = x, y = mean, colour = cover)) +
+  scale_colour_manual(name = "", breaks = c("All"), values = c("black")) +
+  xlab("Slope, rise over run") + ylab("Speed, km/h") +
+  theme_minimal()
+```
 

### 2024-12-11 paper/paper.qmd
latest version of paper

@@
 ```{r}
+#| output: false
+
 library(ggplot2)
@@
 library(ggnewscale)
-# library(whitebox)
 library(ggspatial)
 library(minpack.lm)
+
+home <- "/Users/david/Documents/work/mwlr-tpm-antarctica/antarctica"
+basename           <- "dry-valleys"
+dem_resolution     <- "32m" # 10m also available, but probably not critical
+resolution         <- 100   # 100 is high res for rapid iterating: use 200 while testing
+impassable_geology <- "ice"
+
+data_folder        <- str_glue("{home}/_data")
+dem_folder         <- str_glue("{data_folder}/dem")
+
+coastline          <- st_read(str_glue("{data_folder}/coastline.gpkg"))
+basins             <- st_read(str_glue("{dem_folder}/antarctic-basins.gpkg"))
+study_basins       <- st_read(str_glue("{dem_folder}/dry-valleys-and-skelton.gpkg"))
+study_area         <- st_read(str_glue("{dem_folder}/study-area-extent.gpkg"))
+gps_data           <- st_read(str_glue("{data_folder}/cleaned-gps-data/all-gps-traces.gpkg"))
+tiles              <- st_read(str_glue("{dem_folder}/REMA-index-included-tiles.gpkg"))
+shade              <- rast(str_glue("{dem_folder}/{dem_resolution}/dry-valleys-combined-{dem_resolution}-browse.tif"))
+shade8             <- shade |>
+  terra::aggregate(8) |>
+  as.data.frame(xy = TRUE) |>
+  rename(shade = `15_35_32m_v2.0_browse`)
+geology            <- st_read(str_glue("{data_folder}/ata-scar-geomap-geology-v2022-08-clipped.gpkg"))
+contiguous_geologies <- st_read(str_glue("{data_folder}/{basename}/common-inputs/contiguous-geologies.gpkg"))
+
+base_folder        <- str_glue("{data_folder}/{basename}")
+dem_file           <- str_glue("{dem_folder}/{dem_resolution}/{basename}-combined-{dem_resolution}-dem-clipped.tif")
+
+source(str_glue("{home}/paper/raster-to-graph-functions-paper.R"))
 ```
@@
 ## Our approach and related work
-
 We chose to explore the question of where human impacts are likely to be strongest using an approach closely related to work on patterns of human movement in archaeology [@Verhagen2019] where likely and potential movement paths of humans across landscapes have been used to infer the settlement structure and human geography of large-scale landscapes.
@@
 ### Hiking functions {#sec-hiking-functions}
-
 @Prisner2023 provide an overview of a variety of functions that have been used to model how hiking times and speeds vary with terrain slope. 
@@
 
-get_graph_adjacencies <- function(G) {
-  edges <- ends(G, E(G))
-  list(i = edges[, 1], j = edges[, 2])
-}
-
-make_linestring <- function(x1, y1, x2, y2) {
-  st_linestring(matrix(c(x1, y1, x2, y2), ncol = 2, byrow = TRUE))
-}
-
-get_graph_as_line_layer <- function(G, crs) {
-  ij <- get_graph_adjacencies(G)
-  i <- ij$i
-  j <- ij$j
-  edges_as_sf <- mapply(make_linestring, 
-                        V(G)$x[i], V(G)$y[i], V(G)$x[j], V(G)$y[j],
-                        SIMPLIFY = FALSE) |>
-    st_sfc() |>
-    st_as_sf(crs = crs) |>
-    rename(geometry = x)
-  edges_as_sf                      
-}
-
 get_lattice <- function(dist, pts) {
@@
 The Skelton basin was included because while the expedition GPS data was ostensibly collected in the McMurdo Dry Valleys, it actually extends into that basin as shown in @fig-study-area-map.
-Elevation data from the Reference Elevation Model of Antarctica project @Howat2022, and geology from GeoMAP @Cox2023 are shown in @fig-dem-and-geology.
+Elevation data from the Reference Elevation Model of Antarctica (REMA) project @Howat2022, and geology from GeoMAP @Cox2023 are shown in @fig-dem-and-geology.
 The five largest areas of contiguous non-ice surface geology across the study area shown in @fig-five-subregions were chosen to be the specific sites for more detailed exploration using the methods set out in this paper.
 These range in size from around 320 to 2600 square kilometres.
-
-```{r}
-#| output: false
-data_folder <- "/Users/david/Documents/work/mwlr-tpm-antarctica/antarctica/_data"
-dem_folder  <- "/Users/david/Documents/work/mwlr-tpm-antarctica/antarctica/_data/dem"
-
-coastline <- st_read(str_glue("{data_folder}/coastline.gpkg"))
-basins <- st_read(str_glue("{dem_folder}/antarctic-basins.gpkg"))
-study_basins <- st_read(str_glue("{dem_folder}/dry-valleys-and-skelton.gpkg"))
-study_area <- st_read(str_glue("{dem_folder}/study-area-extent.gpkg"))
-gps_data <- st_read(str_glue("{data_folder}/cleaned-gps-data/all-gps-traces.gpkg"))
-tiles <- st_read(str_glue("{dem_folder}/REMA-index-included-tiles.gpkg"))
-shade <- rast(str_glue("{dem_folder}/32m/dry-valleys-combined-32m-browse.tif"))
-shade8 <- shade |>
-  terra::aggregate(8) |>
-  as.data.frame(xy = TRUE) |>
-  rename(shade = `15_35_32m_v2.0_browse`)
-geology <- st_read(str_glue("{data_folder}/ata-scar-geomap-geology-v2022-08-clipped.gpkg"))
-contiguous_geologies <- st_read(str_glue("{data_folder}/dry-valleys/common-inputs/contiguous-geologies.gpkg"))
-```
 
@@
 # gps data and its bbox
-gps_data <- st_read(str_glue("{data_folder}/cleaned-gps-data/all-gps-traces.gpkg"))
 bb <- gps_data |> st_bbox() # this is to allow plotting restricted to GPS extents
@@
 # geologies data
-geologies <- 
-  st_read(str_glue("{data_folder}/ata-scar-geomap-geology-v2022-08-clipped.gpkg")) |> 
+geologies <- geology |> 
   dplyr::select(POLYGTYPE) |>
@@
 
-GPS data and crudely estimated hiking functions before and after filtering the to remove fixes associated iwht non-purposive movement.
+GPS data and crudely estimated hiking functions before and after filtering the to remove fixes associated with non-purposive movement.
 :::
@@
 
-# see https://rmazing.wordpress.com/2013/08/14/predictnls-part-1-monte-carlo-simulation-confidence-intervals-for-nls-models/
-predictNLS <- function(object, newdata, level = 0.95, nsim = 10000, ...) {
-  require(MASS, quietly = TRUE)
-  ## get right-hand side of formula
-  RHS <- as.list(object$call$formula)[[3]]
-  EXPR <- as.expression(RHS)
-  ## all variables in model
-  VARS <- all.vars(EXPR)
-  ## coefficients
-  COEF <- coef(object)
-  ## extract predictor variable    
-  predNAME <- setdiff(VARS, names(COEF))  
-  ## take fitted values, if 'newdata' is missing
-  if (missing(newdata)) {
-    newdata <- eval(object$data)[predNAME]
-    colnames(newdata) <- predNAME
-  }
-  ## check that 'newdata' has same name as predVAR
-  if (names(newdata)[1] != predNAME) stop("newdata should have name '", predNAME, "'!")
-  ## get parameter coefficients
-  COEF <- coef(object)
-  ## get variance-covariance matrix
-  VCOV <- vcov(object)
-  ## augment variance-covariance matrix for 'mvrnorm' 
-  ## by adding a column/row for 'error in x'
-  NCOL <- ncol(VCOV)
-  ADD1 <- c(rep(0, NCOL))
-  ADD1 <- matrix(ADD1, ncol = 1)
-  colnames(ADD1) <- predNAME
-  VCOV <- cbind(VCOV, ADD1)
-  ADD2 <- c(rep(0, NCOL + 1))
-  ADD2 <- matrix(ADD2, nrow = 1)
-  rownames(ADD2) <- predNAME
-  VCOV <- rbind(VCOV, ADD2) 
-  ## iterate over all entries in 'newdata' as in usual 'predict.' functions
-  NR <- nrow(newdata)
-  respVEC <- numeric(NR)
-  seVEC <- numeric(NR)
-  varPLACE <- ncol(VCOV)   
-  ## define counter function
-  counter <- function (i) {
-    if (i%%10 == 0) 
-      cat(i)
-    else cat(".")
-    if (i%%50 == 0) 
-      cat("\n")
-    flush.console()
-  }
-  outMAT <- NULL 
-  for (i in 1:NR) {
-    counter(i)
-    ## get predictor values and optional errors
-    predVAL <- newdata[i, 1]
-    if (ncol(newdata) == 2) predERROR <- newdata[i, 2] else predERROR <- 0
-    names(predVAL) <- predNAME  
-    names(predERROR) <- predNAME  
-    ## create mean vector for 'mvrnorm'
-    MU <- c(COEF, predVAL)
-    ## create variance-covariance matrix for 'mvrnorm'
-    ## by putting error^2 in lower-right position of VCOV
-    newVCOV <- VCOV
-    newVCOV[varPLACE, varPLACE] <- predERROR^2
-    ## create MC simulation matrix
-    simMAT <- mvrnorm(n = nsim, mu = MU, Sigma = newVCOV, empirical = TRUE)
-    ## evaluate expression on rows of simMAT
-    EVAL <- try(eval(EXPR, envir = as.data.frame(simMAT)), silent = TRUE)
-    if (inherits(EVAL, "try-error")) stop("There was an error evaluating the simulations!")
-    ## collect statistics
-    PRED <- data.frame(predVAL)
-    colnames(PRED) <- predNAME   
-    FITTED <- predict(object, newdata = data.frame(PRED))
-    MEAN.sim <- mean(EVAL, na.rm = TRUE)
-    SD.sim <- sd(EVAL, na.rm = TRUE)
-    MEDIAN.sim <- median(EVAL, na.rm = TRUE)
-    MAD.sim <- mad(EVAL, na.rm = TRUE)
-    QUANT <- quantile(EVAL, c((1 - level)/2, level + (1 - level)/2))
-    RES <- c(FITTED, MEAN.sim, SD.sim, MEDIAN.sim, MAD.sim, QUANT[1], QUANT[2])
-    outMAT <- rbind(outMAT, RES)
-  }
-  colnames(outMAT) <- c("fit", "mean", "sd", "median", "mad", names(QUANT[1]), names(QUANT[2]))
-  rownames(outMAT) <- NULL
-  cat("\n")
-  return(outMAT)  
-}
-
 conf <- 0.95
@@
 ggplot() +
-  geom_ribbon(data = predictions |> filter(cover != "All"), 
-              aes(x = x, ymin = `2.5%`, ymax = `97.5%`, group = cover, fill = cover), 
-              alpha = 0.35, linewidth = 0) + 
-  scale_fill_brewer(palette = "Set1", name = "Cover") +
   geom_ribbon(data = predictions |> filter(cover == "All"), 
@@
               colour = "black", lty = "dashed", fill = "#00000000", linewidth = 0.35) +
-  geom_line(data = predictions |> filter(cover == "All"), aes(x = x, y = mean, colour = cover)) +
+  # geom_line(data = predictions |> filter(cover == "All"), aes(x = x, y = mean, colour = cover)) +
   scale_colour_manual(name = "", breaks = c("All"), values = c("black")) +
+  geom_ribbon(data = predictions, 
+              aes(x = x, ymin = `2.5%`, ymax = `97.5%`, group = cover, fill = cover), 
+              alpha = 0.35, linewidth = 0) + 
+  scale_fill_manual(values = c("black", "firebrick", "dodgerblue"), name = "Cover") +
   xlab("Slope, rise over run") + ylab("Speed, km/h") +
@@
 
+The estimatedhiking functions associated with the two landcovers are
+$$
+\begin{array}{rcl}
+v_{\mathrm{moraine}} & = & 4.17\,\exp\left[{-\frac{(\theta+0.0526)^2}{0.236}}\right] \\
+v_{\mathrm{rock}}    & = & 3.76\,\exp\left[{-\frac{(\theta+0.119)^2}{0.365}}\right]
+\end{array}
+$$ {#eq-estimated-hiking-functions}
+
+where the different maximum speeds (in kilometres per hour) and slopes of maximum speed are apparent.
+
 ### Landscapes as graphs
-We developed R code [@RCoreTeam2024] to build graphs (i.e.
-networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions.
-Graphs are stored as igraph package [@csardi2006igraph; @Csardi2024] graph objects for further analysis.
+We developed R code [@RCoreTeam2024] to build graphs (i.e. networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions. 
+Graphs are manipulated as igraph package [@csardi2006igraph; @Csardi2024] graph objects for further analysis.
+An important decision in constructing graphs is choice of the spacing of the hexagonal lattice, and also of the underlying DEM from which graph vertex elevations are derived.
+Given the extent of the study sites (see @fig-five-subregions) it was decided that a hexagonal lattice (see @fig-lattice-adjacencies) with hexagons equivalent in area to 100 metre square cells as appropriate.
+The hexagon centre to centre spacing of this lattice is given by 
+$$
+100\sqrt{\frac{2}{\sqrt{3}}}\approxeq 107.5\,\mathrm{metres}
+$$ {#eq-hex-lattice-resolution}
+Given this lattice resolution we interpolated vertex elevations from a 32m resolution DEM from the REMA project [@Howat2022] by bilinear interpolation using the R `terra` package [@Hijmans2024].
+It would be straightforward to derive vertex elevations from a more detailed DEM if required.
 
-How terrain is handled in graph edge cost estimation (i.e., assigning half the cost from the terrain at the vertex at each end of the edge).
+Edge weights (i.e.
+estimated traversal costs) are assigned by calculating the slope of each directed edge, the estimated hiking speed for that slope, and thus finding how long it should take for an edge to be traversed.
+If the estimated traversal time of an edge in the nominal 100 metre lattice is greater than 30 minutes then it is removed from the graph, along with its 'twin' edge in the opposite direction.
+Removing edges in both directions is partly a practical matter as it simplifies the operation of many graph algorithms, but can also be justified on the basis that a slope steep enough to be a barrier to ascent is unlikely to be traversed when descending.
 
-`igraph` implements betweenness centrality measures [@Freeman1978; @Brandes2001], which we used to identify vertices and/or edges most likely to be traversed by minimum travel time routes across the terrain.
+After removal of all such edges only the largest connected graph component is retained so that the resulting hiking network representation is fully connected with no isolated vertices unreachable from elsewhere in the network remaining.
+A map of the fifth largest study area's hiking network is shown in @fig-hiking-network-example.
+This network includes 30,697 vertices and 174,798 directed edges. 
+The largest of the five study sites (see @fig-five-subregions) results in a network containing almost a quarter of a million vertices and over 1.4 million directed edges.
+A hiking network can be used to explore many connectivity properties of the environment.
+For example, for a chosen origin point, a _shortest path tree_ can be derived showing the route (@fig-shortest-path-tree).
 
-### Betweenness centrality limited by radius.
+```{r}
+#| output: false
+#| label: create-lattice-points
+contiguous_geology <- "05"  # small one for testing
+extent_file        <- str_glue("{base_folder}/contiguous-geologies/{basename}-extent-{contiguous_geology}.gpkg")
+extent             <- st_read(extent_file)
+
+terrain <- rast(dem_file) |>
+  crop(extent |> as("SpatVector")) |>
+  mask(extent |> as("SpatVector"))
+# and for visualization
+shade <- get_hillshade(terrain) |>
+  as.data.frame(xy = TRUE) # this is for plotting in ggplot
+
+# calculate hexagon spacing of equivalent area
+hex_cell_spacing <- resolution * sqrt(2 / sqrt(3))
+
+xy  <- extent |>
+  st_make_grid(cellsize = hex_cell_spacing, what = "centers", square = FALSE) |>
+  st_as_sf() |>
+  rename(geometry = x) |>   # shouldsee https://github.com/r-spatial/sf/issues/2429
+  st_set_crs(st_crs(extent))
+coords <- xy |> st_coordinates()
+
+xy <- xy |>
+  mutate(x = coords[, 1], y = coords[, 2])
+
+z <- terrain |> 
+  terra::extract(xy |> as("SpatVector"), method = "bilinear")
+
+pts <- xy |>
+  mutate(z = z[, 2])
+
+# remove any NAs that might result from interpolation or impassable terrain
+# NAs tend to occur around the edge of study area
+cover <- geology |>
+  dplyr::select(POLYGTYPE) |> 
+  rename(terrain = POLYGTYPE)
+
+pts <- pts |>
+  st_join(cover) |>
+  filter(!is.na(z), terrain != impassable_geology)
+```
+
+```{r}
+G <- graph_from_points(pts, hex_cell_spacing * 1.1)
+G <- assign_movement_variables_to_graph_2(G, xyz = pts, terrain = pts$terrain)
+V(G)$id <- 1:length(G)
+
+# now simplify by removing all edges that have est. costs > 30 min for 100m res edges
+# scaled appropriately i.e. if resolution is coarser then that cutoff should be longer
+# and extracting the strong largest component (connected both directions)
+cost_cutoff <- 0.5 * resolution / 100
+G <- delete_edges(G, E(G)[which(E(G)$cost > cost_cutoff)]) |>
+  largest_component(mode = "strong")
+
+remaining_nodes <- V(G)$id
+pts <- pts |> slice(remaining_nodes)
+```
+
+::: {#fig-hiking-network-example layout="[[50,-1,50], [50,-1,50]]"}
+
+```{r}
+#| label: fig-map-of-hiking-network
+#| fig-cap: "Map of a hiking network coloured by estimated traversal times of edges" 
+bb <- (extent |> st_bbox() |> as.vector()) + c(1.05e4, 1.25e4, -1.45e4, -1.35e4)
+ggplot() +
+  geom_raster(data = shade, aes(x = x, y = y, fill = hillshade), alpha = 0.65) +
+  scale_fill_distiller(palette = "Greys", direction = 1) +
+  geom_sf(data = G |> get_graph_as_line_layer(append_attributes = TRUE),
+          aes(colour = cost), linewidth = 0.05) +
+  scale_colour_viridis_c(option = "A", direction = -1, name = "Cost, hours") +
+  coord_sf() +
+  guides(fill = "none") +
+  annotate("rect", xmin = bb[1], ymin = bb[2], xmax = bb[3], ymax = bb[4], 
+           fill = NA, colour = "red", linewidth = 0.35) +
+  annotation_scale(plot_unit = "m", height = unit(0.1, "cm")) +
+  theme_void()
+```
+
+```{r}
+#| label: fig-zoomed-in-map-of-hiking-network
+#| fig-cap: "Red-outlined area zoomed in revealing edges removed from graph due to steep slopes" 
+ggplot() +
+  geom_raster(data = shade, aes(x = x, y = y, fill = hillshade), alpha = 0.65) +
+  scale_fill_distiller(palette = "Greys", direction = 1) +
+  geom_sf(data = G |> get_graph_as_line_layer(append_attributes = TRUE),
+          aes(colour = cost), linewidth = 0.2) +
+  scale_colour_viridis_c(option = "A", direction = -1, name = "Cost, hours") +
+  coord_sf(xlim = bb[c(1, 3)], ylim = bb[c(2, 4)], expand = FALSE) +
+  guides(fill = "none", colour = "none") +
+  theme_void()
+```
+
+```{r}
+#| label: fig-shortest-path-tree
+#| fig-cap: "Shortest path tree of a hiking network from the indicated origin" 
+centroid <- extent |>
+  st_centroid() |> 
+  st_nearest_feature(pts)
+
+spt <- G |> get_shortest_path_tree(vid = centroid) |>
+  get_graph_as_line_layer()
+
+bb <- (extent |> st_bbox() |> as.vector()) +  + c(1.35e4, 0.8e4, -0.9e4, -1.55e4)
+ggplot() +
+  geom_sf(data = extent, colour = NA) +
+  geom_sf(data = spt, colour = "black", linewidth = 0.05) +
+  geom_sf(data = pts |> slice(centroid)) +
+  coord_sf() +
+  guides(fill = "none", colour = "none") +
+  annotate("rect", xmin = bb[1], ymin = bb[2], xmax = bb[3], ymax = bb[4], 
+           fill = NA, colour = "red", linewidth = 0.35) +
+  theme_void()
+```
+
+```{r}
+#| label: fig-zoomed-in-shortest-path-tree
+#| fig-cap: "Red-outlined area zoomed in view of the shortest path tree" 
+ggplot() +
+  geom_raster(data = shade, aes(x = x, y = y, fill = hillshade), alpha = 0.65) +
+  scale_fill_distiller(palette = "Greys", direction = 1) +
+  geom_sf(data = spt, colour = "black", linewidth = 0.2) +
+  geom_sf(data = pts |> slice(centroid)) +
+  coord_sf(xlim = bb[c(1, 3)], ylim = bb[c(2, 4)], expand = FALSE) +
+  guides(fill = "none", colour = "none") +
+  theme_void()
+```
+
+A hiking network derived for the fifth largest study site shown in @fig-five-subregions.
+
+:::
+
+### Betweenness centrality limited by radius
+A particular graph connectivity property we have used to reveal the relative likelihood of different parts of a terrain being frequently traversed is _betweenness centrality_.
+In a graph $G=(V,E)$ a path $P$ is an ordered sequence of vertices:
+$$
+P=\left(v_1,v_2,\ldots,v_n\right)
+$$ {$eq-path}
+such that each consecutive pair of vertices $v_i$ and $v_{i+1}$ is adjacent, i.e. connected by a directed edge $e_{i,j}$.
+The simple length of a path is the number of edges it contains.
+The weighted length of a patch is the sum over its edges of a _weight_ or _cost_ associated with each edge, which we denote $w_{i,j}$, that is the length $L$ of a path $P$ is given by
+$$
+L(P)=\sum_{i=1}^{n-1} w_{i,i+1}
+$$ {#eq-path-length}
+The _shortest path_ from $v_i$ to $v_j$ is the path $P=\left(v_i,\dots,v_*,\ldots,v_j\right)$ such that $L(P)$ is minimised.
+In a regular lattice such as our hiking networks there are many equal length shortest paths between any pair of vertices.
+When we consider edge weights, then the shortest path will be unique, or one of only a small number of possibilities of equal length.
+
+Shortest paths are used to develop many measures of graph centrality [@Freeman1978].
+For example, the mean shortest path length of a vertex is the mean of all the shortest path lengths from that vertex to every other vertex in the graph.
+A vertex with low mean shortest path length might be considered more central than another a vertex with a higher mean shortest path length.
+
+One measure of vertex centrality in a graph is _betweenness centrality_. 
+The betweenness centrality of a vertex is the total number of times a vertex is on shortest paths between every other pair of vertices in the network.
+When the number of shortest paths or _geodesics_ between two vertices $v_j$ and $v_k$ is $g_{jk}$, each appearance on those shortest paths of a vertex only contributes $1/g_{jk}$ to that vertex's betweenness centrality.
+Formally, if we denote the number of times $v_i$ appears on shortest paths between $v_j$ and $v_k$ by $g_{jk}(v_i)$, then we can write this as
+$$
+\mathrm{betweenness}(v_i)=\sum_{k=1}^n\sum_{j=1}^n\frac{g_{jk}(v_i)}{g_{jk}}\forall i\neq j\neq k
+$$ {#eq-vertex-betweenness}
+Betweenness centrality is particularly relevant to applications where we are interested in graph vertices that control movement through the graph.
+An edge betweenness centrality measure can also be calculated on similar principles, but is not considered further here.
+
+As might be expected betweenness centralities are computationally demanding to calculate.
+The time complexity of early algorithms was $\mathcal{O}(n^3)$ where $n$ is the number of vertices in the graph [@Brandes2001].
+An implementation of Brandes's algorithm [-@Brandes2001] which has time complexity $\mathcal{O}(nm + n^2\log n)$ where $m$ is the number of edges in the graph, is provided in the `igraph` package [@Csardi2024].
+Even with this improvement in performance, in our application such computational complexity is a strong motivation for working at a nominal 100 metre resolution.
+Halving the resolution to 50 metres would increase the number of graph vertices four-fold, and lead to a substantial increases in the times taken to calculate betweenness centralities.
+
+Fortunately, the `igraph` implementation of betweennees centrality provides an option to _radius-limit_ between calculations, meaning that only paths shorter than a specified radius are considered in the counting of appearances on shortest paths.
+While 'all graph' betweeness centralities can be standardised relative to a theoretical maximum of $(n-1)(n-2)$, not straightforward standardisation is possible for the radius-limited case.
+Since we are primarily interested in betweenness centrality as a measure of the _relative_ vulnerability of different locations to human impacts, we linearly rescale measured betweennees measures with respect to the results of a given analysis for visualisation purposes.
 

### 2024-12-12 paper/paper.qmd
latest version of paper

@@
 lightbox: true
+navbar: false
+sidebar: false
 format: 
@@
 library(cols4all)
+library(ggnewscale)
+library(ggspatial)
+
 library(tidyr)
 library(dplyr)
+library(purrr)
+library(stringr)
+
 library(igraph)
+
 library(sf)
 library(terra)
+library(spatstat)
 library(akima)
-library(stringr)
-library(ggnewscale)
-library(ggspatial)
+
 library(minpack.lm)
@@
 
-### Betweenness centrality limited by radius
-A particular graph connectivity property we have used to reveal the relative likelihood of different parts of a terrain being frequently traversed is _betweenness centrality_.
-In a graph $G=(V,E)$ a path $P$ is an ordered sequence of vertices:
-$$
-P=\left(v_1,v_2,\ldots,v_n\right)
-$$ {$eq-path}
-such that each consecutive pair of vertices $v_i$ and $v_{i+1}$ is adjacent, i.e. connected by a directed edge $e_{i,j}$.
-The simple length of a path is the number of edges it contains.
-The weighted length of a patch is the sum over its edges of a _weight_ or _cost_ associated with each edge, which we denote $w_{i,j}$, that is the length $L$ of a path $P$ is given by
+### Betweenness centrality
+The connectivity properties of a hiking graph can be described using a range of measures of graph structure and used to reveal the relative likelihood of different parts of a terrain being frequently traversed.
+The particular structural measure we focus on is _betweenness centrality_, which is explained below.
+
+In a graph $G=(V,E)$ a path $P$ is an ordered sequence of vertices, $P=\left(v_1,v_2,\ldots,v_n\right)$ such that each consecutive pair of vertices $v_i$ and $v_{i+1}$ is adjacent, i.e.
+connected by a directed edge $e_{i,j}$.
+The _length_ of a path is the number of edges it contains.
+The _weighted length_ or _cost_ of a path is the sum over its edges of a _weight_ or _cost_ associated with each edge, which we here denote $w_{i,j}$, i.e., the length $L$ of a path $P$ is given by
 $$
@@
 $$ {#eq-path-length}
-The _shortest path_ from $v_i$ to $v_j$ is the path $P=\left(v_i,\dots,v_*,\ldots,v_j\right)$ such that $L(P)$ is minimised.
-In a regular lattice such as our hiking networks there are many equal length shortest paths between any pair of vertices.
-When we consider edge weights, then the shortest path will be unique, or one of only a small number of possibilities of equal length.
 
-Shortest paths are used to develop many measures of graph centrality [@Freeman1978].
-For example, the mean shortest path length of a vertex is the mean of all the shortest path lengths from that vertex to every other vertex in the graph.
-A vertex with low mean shortest path length might be considered more central than another a vertex with a higher mean shortest path length.
+The _shortest path_ from $v_i$ to $v_j$ is the path $P=\left(v_i,\dots,v_*,\ldots,v_j\right)$ starting at$v_i$ and ending at $v_i$ passing through intervening vertices $\left(v_*\right)$ such that $L(P)$ is minimised.
+In a regular lattice such as our hiking networks there are many shortest paths of equal length if the weight associated with each edge is equal, as it would be if we based it solely on the length of the edge between each pair of vertices.
+When we consider edge traversal costs derived from the slope of each edge and a hiking function, then the shortest path will be unique, or one of only a small number of possibilities of equal total cost.
 
-One measure of vertex centrality in a graph is _betweenness centrality_. 
-The betweenness centrality of a vertex is the total number of times a vertex is on shortest paths between every other pair of vertices in the network.
-When the number of shortest paths or _geodesics_ between two vertices $v_j$ and $v_k$ is $g_{jk}$, each appearance on those shortest paths of a vertex only contributes $1/g_{jk}$ to that vertex's betweenness centrality.
-Formally, if we denote the number of times $v_i$ appears on shortest paths between $v_j$ and $v_k$ by $g_{jk}(v_i)$, then we can write this as
+Shortest paths are used to develop many measures of vertext centrality in graphs [@Freeman1978].
+One such measure of vertex centrality is _betweenness centrality_. 
+The betweenness centrality of a vertex is the total number of times that vertex is on shortest paths between every other pair of vertices in the network.
+If we denote the number of shortest paths or _geodesics_ between two vertices $v_j$ and $v_k$ by $g_{jk}$, then each appearance of a vertex $v_i$ on the shortest path between those vertices only contributes $1/g_{jk}$ to the betweenness centrality of $v_i$.
+Formally, if we denote the number of times vertex $v_i$ appears on shortest paths between $v_j$ and $v_k$ by $g_{jk}(v_i)$, then its betweenness $b_i$ is given by
 $$
-\mathrm{betweenness}(v_i)=\sum_{k=1}^n\sum_{j=1}^n\frac{g_{jk}(v_i)}{g_{jk}}\forall i\neq j\neq k
+b_i=\sum_{k=1}^n\sum_{j=1}^n\frac{g_{jk}(v_i)}{g_{jk}}\forall i\neq j\neq k
 $$ {#eq-vertex-betweenness}
-Betweenness centrality is particularly relevant to applications where we are interested in graph vertices that control movement through the graph.
-An edge betweenness centrality measure can also be calculated on similar principles, but is not considered further here.
+Betweenness centrality is particularly relevant to applications where we are interested in how important vertices are to movement across the graph.
+An edge betweenness centrality measure can also be calculated on similar principles, but is not considered further here, in part because visualization is challenging given the potential for different scores for the two directed edges between each pair of vertices.
+Vertex betweenness on the other hand yields a single value for each vertex.
 
 As might be expected betweenness centralities are computationally demanding to calculate.
-The time complexity of early algorithms was $\mathcal{O}(n^3)$ where $n$ is the number of vertices in the graph [@Brandes2001].
-An implementation of Brandes's algorithm [-@Brandes2001] which has time complexity $\mathcal{O}(nm + n^2\log n)$ where $m$ is the number of edges in the graph, is provided in the `igraph` package [@Csardi2024].
+The time complexity of early algorithms was $\mathcal{O}(n^3)$, where $n$ is the number of vertices in the graph [@Brandes2001].
+An implementation of Brandes's algorithm [-@Brandes2001] with time complexity $\mathcal{O}(nm + n^2\log n)$, where $m$ is the number of edges in the graph, is provided in the `igraph` package [@Csardi2024].
 Even with this improvement in performance, in our application such computational complexity is a strong motivation for working at a nominal 100 metre resolution.
@@
 
-Fortunately, the `igraph` implementation of betweennees centrality provides an option to _radius-limit_ between calculations, meaning that only paths shorter than a specified radius are considered in the counting of appearances on shortest paths.
-While 'all graph' betweeness centralities can be standardised relative to a theoretical maximum of $(n-1)(n-2)$, not straightforward standardisation is possible for the radius-limited case.
-Since we are primarily interested in betweenness centrality as a measure of the _relative_ vulnerability of different locations to human impacts, we linearly rescale measured betweennees measures with respect to the results of a given analysis for visualisation purposes.
+Results of betweenness centrality for our example hiking network are shown in @fig-betweenness-map.
+'All-paths' betweenness centralities can be standardised relative to a theoretical maximum of $(n-1)(n-2)$, no straightforward standardisation is possible for the radius-limited case.
+Since we are primarily interested in betweenness as a measure of the _relative_ vulnerability of different locations to human impacts, we linearly rescaled betweenness scores in all cases with respect to the range of scores across the graph being analysed.
+The rescaled betweenness, $b_i^\prime$ for vertex $v_i$ is given by
+$$
+b_i^\prime=\frac{b_i-b_{\min}}{b_{\max}-b_{\min}}
+$$ {#eq-betweenness-rescaling}
+This approach is also applicable to radius-limited betweenness scores (see @sec-radius-limited-betweenness).
+
+When all-paths betweenness is calculated, bottlenecks between large sub-areas are strongly highlighted as is the case for the 'inner bend' of the relatively narrow neck of navigable terrain that connects the east and west sub-areas of this site.
+The other feature of this map is the identification of a distinctive structure of 'arterial' routes.
+
+```{r}
+if (!"betweenness" %in% vertex_attr_names(G)) {
+  V(G)$betweenness <- betweenness(G, weights = E(G)$cost, cutoff = -1) |>
+    scales::rescale(to = c(0, 100))
+}
+```
+
+```{r}
+#| label: fig-betweenness-map
+#| fig-cap: "Relative vertex betweenness of vertices in the graph from @fig-hiking-network-example."
+G_sf <- G |> get_graph_as_point_layer() |>
+  st_buffer(hex_cell_spacing / 2, nQuadSegs = 2)
+ggplot(G_sf) +
+  geom_sf(aes(fill = betweenness), colour = "NA") +
+  scale_fill_viridis_c(option = "A", direction = -1, name = "Relative\nbetweenness") +
+  theme_void()
+```
+
+### Betweenness centrality limited by radius {#sec-radius-limited-betweenness}
+The `igraph` implementation of betweenness centrality provides an option to _radius limit_ betweenness calculations, meaning that only paths shorter than a specified radius (expressed in cost units, here of time) are considered in counting the appearance of vertices on shortest paths.
+This approach is particularly useful for large graphs where the time complexity of calculating radius limited betweenness scales more favourably than cited above, because the number of shortest paths that must be identified is greatly reduced.
+We informally confirmed the intuition that for a given radius limit the time take to compute betweenness scales approximately linearly with the number of vertices in the graph, since the number of shortest paths local to each vertex, and on which it might be found is similar.
+
+Results for a series of radius limits are shown in @fig-radius-limited-betweenness.
+The results here are interesting.
+At very short radii (the 30 minutes case) many locally important paths are identified as potentially relatively heavily trafficked.
+As the radius increases the pattern emphasises potential paths that are important across wider areas, eventually tending toward the arterial structure of the all-paths case.
+We discuss these results more fully in @sec-discussion.
+
+```{r}
+radii <- c(0.5, 1, 1.5, 2, 3, 5)
+betweenness_results <- list()
+for (i in seq_along(radii)) {
+  r <- radii[i]
+  b_r <- betweenness(G, weights = E(G)$cost, cutoff = r) |>
+    scales::rescale(to = c(0, 100))
+  betweenness_results[[i]] <- G_sf |>
+    mutate(radius = str_glue("{r*60} mins"), betweenness = b_r)
+}
+```
+
+```{r}
+#| label: fig-radius-limited-betweenness
+#| fig-cap: "Vertex betweenness maps based on different radius limits. As in @fig-betweenness-map the visualization colouring is based on linear scaling of raw betweenness scores."
+betweenness_results_df <- betweenness_results |>
+  bind_rows() |>
+  mutate(radius = factor(radius, levels = str_glue("{radii * 60} mins")
+))
+
+ggplot(betweenness_results_df) +
+  geom_sf(aes(fill = betweenness), colour = "NA") +
+  scale_fill_viridis_c(option = "A", direction = -1) +
+  guides(fill = "none") +
+  facet_wrap( ~ radius, ncol = 3) +
+  theme_void()
+```
 
 ### Impact minimizing networks
+In this section we explore developing the betweenness centrality for a limited set of locations, which might represent a base camp, rest sites, and experimental sites for a season's expedition.
 
-Tentative proposal for impact minimizing networks based on minimum spanning trees, but noting the issue with respect to directed graphs when these would more correctly be arborescences [@Korte2018a].
+This is relatively straightforward and like radius limited betweenness involves limiting the shortest paths considered to only those among a chosen subset of vertices in the graph.
+One slight simplification we make for performance reasons is to weight every appearance of a vertex on a geodesic as equal, even if multiple equal length shortest paths exist between two vertices. 
+In practice, this is unlikely to have much effect on the resulting estimates of relative betweenness, since with real-valued edge costs two paths of exactly equal cost are unlikely.
+An example result is shown in @fig-betweenness-limited-sites.
+One refinement shown in the figure is that sites have been waited by relative importance so that a site such as an expedition base camp can be expected to originate more trips than other sites.
 
-## Discussion
+```{r}
+# make a random set of locations with guaranteed minimum separation
+origin_pts <- rSSI(2000, n = 20, win = as.owin(extent)) |> 
+  st_as_sf() |> 
+  st_set_crs(st_crs(extent)) |> 
+  slice(-1) |> 
+  dplyr::select(-label)
 
-Blah
+# get the sequence numbers of thes in the pts data
+origin_is <- origin_pts |>
+  st_nearest_feature(pts)
 
-## Conclusions
+# make up a point dataset of these origins with randomly assigned importance
+origin_vs <- pts |>
+  mutate(id = row_number()) |>
+  slice(origin_is) |>
+  mutate(Importance = sample((seq(2, 50, 2) / 5) ^ 2, n(), 
+                             replace = TRUE, prob = (seq(50, 2, -2) / 5) ^ 2))
 
-Blah
+# put all vertices on a shortest path to begin
+vs_on_shortest_paths <- V(G) |> 
+  as.vector()
+
+# make a data frame we can join the results into
+for (i in seq_along(origin_vs$id)) {
+  # one entry for each appearance on a shortest path
+  SP <- shortest_paths(G, origin_vs$id[i], origin_vs$id[-i], weights = edge_attr(G, "cost"))
+  new_shortest_paths <- SP$vpath |>
+    lapply(as.vector) |>
+    reduce(c) |>
+    rep(origin_vs$Importance[i])
+  vs_on_shortest_paths <- vs_on_shortest_paths |>
+    c(new_shortest_paths)
+}
+shortest_path_counts <- vs_on_shortest_paths |>
+  table() |>
+  as.vector()
+
+Gsp <- set_vertex_attr(G, "sp_betweenness", value = shortest_path_counts - 1)
+G_sf_od <- G |>
+  set_vertex_attr("sp_betweenness", value = shortest_path_counts - 1) |>
+  get_graph_as_point_layer() |>
+  st_buffer(hex_cell_spacing / 2, nQuadSegs = 2) |>
+  filter(sp_betweenness > 0)
+```
+
+```{r}
+#| label: fig-betweenness-limited-sites
+#| fig-cap: The network of shortest paths among a weighted set of expedition sites.
+ggplot() +
+  geom_raster(data = shade, aes(x = x, y = y, fill = hillshade), alpha = 0.65) +
+  scale_fill_distiller(palette = "Greys", direction = 1) +
+  guides(fill = "none") +
+  new_scale_fill() +
+  geom_sf(data = G_sf_od, aes(fill = sp_betweenness), colour = NA) +
+  scale_fill_viridis_c(option = "A", direction = -1) +
+  geom_sf(data = origin_vs, aes(size = Importance), colour = "black", pch = 1) +
+  guides(fill = "none") +
+  theme_void()
+```
+
+A further analytical possibility is to use the hiking network to develop a set of routes among the sites that might minimise total impact, by confining movement to a limited set of paths.
+This approach is consistent with work in ecology and park and wildlife management which suggests that path networks can reduce the impact of human movement in many environments [@Cole1995a;@Cole1995;@Kidd2015;@Marion2016;@Piscova2023;@Tomczyk2023].
+
+The approach taken is first to first determine the lengths of all the shortest paths among all the sites.
+Then a _minimum spanning tree_ for the sites is calculated, which is a graph with minimum total edge length that connects all the sites [@Prim1957].
+One refinement here, which we do not account for is that algorithms for finding the minimum spanning tree of a directed graph (referred to a _minimum-cost arborescence_ or _optimal branching_) are more complex [@Korte2018a], and not at present supported in `igraph`.
+While there is some asymmetry in paths between locations depending on the direction of travel, it is not extreme, and so we approximate a solution by finding the minimum spanning tree links (i.e.
+direct connections) among sites treating our network as undirected, and then determine paths over the ground to make the connections between sites based on the directed hiking network.
+The resulting path network for the same set of sites is shown in @fig-approx-minimum-arborescence.
+
+```{r}
+dist_m <- distances(G, origin_vs$id, origin_vs$id, weights = edge_attr(G, "cost"))
+xs <- vertex_attr(G, "x", V(G)[origin_vs$id])
+ys <- vertex_attr(G, "y", V(G)[origin_vs$id])
+approx_min_arborescence <- dist_m |> 
+  graph_from_adjacency_matrix(mode = "directed", weighted = TRUE) |>
+  set_vertex_attr("x", value = xs) |>
+  set_vertex_attr("y", value = ys) |>
+  mst() # this applies algorithms appropriate to an undirected graph
+```
+
+```{r}
+dir_m <- approx_min_arborescence |> as_adjacency_matrix(sparse = FALSE)
+undir_m <- ((dir_m + t(dir_m)) > 0)
+which_ijs <- which(undir_m, arr.ind = TRUE)
+
+# put all vertices on a shortest path to begin
+vs_on_shortest_paths <- V(G) |> 
+  as.vector()
+
+# make a data frame we can join the results into
+for (k in 1:nrow(which_ijs)) {
+  i <- which_ijs[k, 1]
+  j <- which_ijs[k, 2]
+  # one entry for each appearance on a shortest path
+  SP <- shortest_paths(G, origin_vs$id[i], origin_vs$id[j], weights = edge_attr(G, "cost"))
+  new_shortest_paths <- SP$vpath[[1]] |>
+    as.vector() |>
+    rep(origin_vs$Importance[i])
+  vs_on_shortest_paths <- vs_on_shortest_paths |>
+    c(new_shortest_paths)
+}
+shortest_path_counts <- vs_on_shortest_paths |>
+  table() |>
+  as.vector()
+
+G_sf_min <- G |>
+  set_vertex_attr("sp_betweenness", value = shortest_path_counts - 1) |>
+  get_graph_as_point_layer() |>
+  st_buffer(hex_cell_spacing / 2, nQuadSegs = 2) |>
+  filter(sp_betweenness > 0)
+```
+
+```{r}
+#| label: fig-approx-minimum-arborescence
+#| fig-cap: "A potential impact minimising route network among sites."
+ggplot() +
+  geom_raster(data = shade, aes(x = x, y = y, fill = hillshade), alpha = 0.65) +
+  scale_fill_distiller(palette = "Greys", direction = 1) +
+  guides(fill = "none") +
+  new_scale_fill() +
+  geom_sf(data = G_sf_min, aes(fill = sp_betweenness), colour = NA) +
+  scale_fill_viridis_c(option = "A", direction = -1) +
+  geom_sf(data = origin_vs, aes(size = Importance), colour = "black", pch = 1) +
+  guides(fill = "none", size = "none") +
+  theme_void()
+```
+
+## Discussion and conclusions {#sec-discussion}

### 2024-12-12 paper/paper.qmd
frozen paper prior to mtg with Fraser

@@
 # get the sequence numbers of thes in the pts data
+
 origin_is <- origin_pts |>
@@
   slice(origin_is) |>
-  mutate(Importance = sample((seq(2, 50, 2) / 5) ^ 2, n(), 
-                             replace = TRUE, prob = (seq(50, 2, -2) / 5) ^ 2))
+  # note: important here to avoid Importance values < 1!
+  mutate(Importance = rgeom(n(), 0.1) + 1)
 
@@
 
-## Discussion and conclusions {#sec-discussion}
+## Discussion {#sec-discussion}
+To cover:
++ Figure of the largest study site?
++ Limitations of the method:
+  + Data concerns, esp. opportunistic nature of the GPS data
+  + Would higher resolution network be useful?
+  + Reliance of shortest paths as a guide to behaviour
+  + Some minor issues regarding the difference between directed and undirected graphs
++ Relevance of radius limits on betweenness: the implicit model of movement in this approach (short horizons, no existing paths, 'keep going for half an hour in this general direction...')
++ Question of embedding paths, i.e., locking in more damage to some parts of the landscape vs others
+
+## Conclusions {#sec-conclusions}
+To cover:
++ Novelty of fitting land cover specific hiking functions
++ A method for exploring potential impacts in a novel environment
++ A method for proposing less damaging plans for expeditions
++ Other?

### 2024-12-12 paper/paper.qmd
updated discussion/conclusion bullets

@@
 To cover:
+
 + Figure of the largest study site?
@@
   + Would higher resolution network be useful?
+  + Possible limitations of a regularly spaced lattice: extend to a lattice based on e.g., TIN?
   + Reliance of shortest paths as a guide to behaviour
@@
 To cover:
+
 + Novelty of fitting land cover specific hiking functions

### 2024-12-19 paper/paper.qmd
added Minetti citation
latest minor revisions before summer break

@@
 We use the estimated hiking functions to build weighted directed graphs as a representation of specific environments in Antarctica. 
-From these we estimate using a variety of graph metrics—particularly betweennness centrality—the relative potential for human environmental impacts arising from scientific activities in those environments.
+From these we estimate using a variety of graph metrics—particularly betweenness centrality—the relative potential for human environmental impacts arising from scientific activities in those environments.
 We also suggest a simple approach to planning science expeditions that might allow for reduced impacts in these environments.
@@
 ## Introduction
+**FRASER TO FRAME / CLARIFY MOTIVATION**
 
@@
 However, there is no theoretical basis for the specific functional forms listed in @tbl-hiking-function-equations. 
-More principled approaches might be developed based on the literature on the physiology of human movement, see e.g. **CITATIONS NEEDED HERE**. 
+More principled approaches might be developed based on the literature on the physiology of human movement [see, e.g., @Minetti2002]. 
 In general, approaches based on minimising energy use yield similar results to empirical speed-slope functions, although it is worth noting that they more reliably generate zig-zag or 'switchback' movement behaviour on steep slopes [@Llobera2007].
@@
 We therefore represent the landscape as a directed graph $G(V,E)$ of vertices and directed edges connecting them.
-In this representaiton, graph vertices $v_i$ are locations with associated elevation and land cover. 
+In this representation, graph vertices $v_i$ are locations with associated elevation and land cover. 
 Vertices minimally have spatial coordinates $(x_i,y_i)$, an elevation $z_i$, and a land cover, $C_i$:
@@
 
-The obvious way to derive such a graph from spatial data is to assign each cell in a raster digital elevation model (DEM) to a graph vertex, so that graph vertices are arranged in a regular square grid or lattice, and this approach has been widely adopted.
+The obvious way to derive such a graph from spatial data is to assign each cell in a raster digital elevation model (DEM) to a graph∏ vertex, so that graph vertices are arranged in a regular square grid or lattice, and this approach has been widely adopted.
 **CITATIONS NEEDED HERE**
@@
 
-Overplotting of the hiking functions including an additional model fitted to all the data, confirms that the fitted functions are sufficiently different to retain separate models for each ground cover (see @fig-models-by-cover-compared).
+Plotting both hiking functions along with an additional model fitted to all the data on the same axes confirms that the fitted functions are sufficiently different to retain separate models for each ground cover (see @fig-models-by-cover-compared).
 Plotting both functions in the same graph makes clearer the difference in maximum speed and slope at maximum speed associated with each ground cover.
@@
 
-The estimatedhiking functions associated with the two landcovers are
+The estimated hiking functions associated with the two land covers are
 $$
@@
 We developed R code [@RCoreTeam2024] to build graphs (i.e. networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions. 
-Graphs are manipulated as igraph package [@csardi2006igraph; @Csardi2024] graph objects for further analysis.
+Graphs are manipulated as `igraph` package [@csardi2006igraph; @Csardi2024] graph objects for further analysis.
 An important decision in constructing graphs is choice of the spacing of the hexagonal lattice, and also of the underlying DEM from which graph vertex elevations are derived.
@@
 
-Shortest paths are used to develop many measures of vertext centrality in graphs [@Freeman1978].
+Shortest paths are used to develop many measures of vertex centrality in graphs [@Freeman1978].
 One such measure of vertex centrality is _betweenness centrality_. 
@@
 At very short radii (the 30 minutes case) many locally important paths are identified as potentially relatively heavily trafficked.
-As the radius increases the pattern emphasises potential paths that are important across wider areas, eventually tending toward the arterial structure of the all-paths case.
+As the radius increases the pattern emphasizes potential paths that are important across wider areas, eventually tending toward the arterial structure of the all-paths case.
 We discuss these results more fully in @sec-discussion.
@@
 # make a random set of locations with guaranteed minimum separation
+set.seed(1)
+
 origin_pts <- rSSI(2000, n = 20, win = as.owin(extent)) |> 
@@
 Then a _minimum spanning tree_ for the sites is calculated, which is a graph with minimum total edge length that connects all the sites [@Prim1957].
-One refinement here, which we do not account for is that algorithms for finding the minimum spanning tree of a directed graph (referred to a _minimum-cost arborescence_ or _optimal branching_) are more complex [@Korte2018a], and not at present supported in `igraph`.
+One refinement here, which we do not account for is that algorithms for finding the minimum spanning tree of a directed graph (referred to as a _minimum-cost arborescence_ or _optimal branching_) are more complex [@Korte2018a], and not at present supported in `igraph`.
 While there is some asymmetry in paths between locations depending on the direction of travel, it is not extreme, and so we approximate a solution by finding the minimum spanning tree links (i.e.
-direct connections) among sites treating our network as undirected, and then determine paths over the ground to make the connections between sites based on the directed hiking network.
-The resulting path network for the same set of sites is shown in @fig-approx-minimum-arborescence.
+direct connections) among sites treating our network as undirected, and then determine paths over the ground to make the identified connections between sites based on the directed hiking network.
+The resulting path network for the same set of sites as in @fig-betweenness-limited-sites is shown in @fig-approx-minimum-arborescence.
 

### 2025-06-07 paper/paper.qmd
paper pdf recompiled using typst and patchwork for multipanel figures

@@
 format: 
+  typst:
+    toc: false
+    number-sections: true
+    fig-format: png
+    fig-dpi: 600
   html:
@@
     self-contained: true
-  pdf:
-    documentclass: article
-    classoption: [11pt]
-    toc: false
-    number-sections: true
-    number-depth: 3
-    linestretch: 1.15
+  # pdf:
+  #   documentclass: article
+  #   classoption: [11pt]
+  #   toc: false
+  #   number-sections: true
+  #   number-depth: 3
+  #   linestretch: 1.15
 execute:
@@
 library(ggplot2)
-library(ggpubr)
 library(grid)
@@
 library(ggspatial)
+library(patchwork)
 
@@
 }
-ggarrange(plotlist = ggplots, ncol = 2, nrow = 2)
+
+wrap_plots(ggplots, nrow = 2)
 ```
@@
 Geospatial data for Antarctica were obtained from sources referenced in @Cox2023, @Cox2023a, and @Felden2023.
-The study area was defined to be the Skelton and Dry Valleys basins, as defined by the NASA Making Earth System Data Records for Use in Research Environments (MEaSUREs) project [@Mouginot2017] and shown in @fig-locator-map.
-The Skelton basin was included because while the expedition GPS data was ostensibly collected in the McMurdo Dry Valleys, it actually extends into that basin as shown in @fig-study-area-map.
-Elevation data from the Reference Elevation Model of Antarctica (REMA) project @Howat2022, and geology from GeoMAP @Cox2023 are shown in @fig-dem-and-geology.
-The five largest areas of contiguous non-ice surface geology across the study area shown in @fig-five-subregions were chosen to be the specific sites for more detailed exploration using the methods set out in this paper.
+The study area was defined to be the Skelton and Dry Valleys basins, as defined by the NASA Making Earth System Data Records for Use in Research Environments (MEaSUREs) project [@Mouginot2017] and shown in @fig-study-area(a).
+The Skelton basin was included because while the expedition GPS data was ostensibly collected in the McMurdo Dry Valleys, it actually extends into that basin as shown in @fig-study-area(b).
+Elevation data from the Reference Elevation Model of Antarctica (REMA) project @Howat2022, and geology from GeoMAP @Cox2023 are shown in @fig-study-area(c).
+The five largest areas of contiguous non-ice surface geology across the study area shown in @fig-study-area(d) were chosen to be the specific sites for more detailed exploration using the methods set out in this paper.
 These range in size from around 320 to 2600 square kilometres.
 
-::: {#fig-study-area layout="[[50,-1,50], [50,-1,50]]"}
-
-```{r}
-#| label: fig-locator-map
-#| fig-cap: "Study area location in Antarctica"
+```{r}
+#| echo: false
+
 theme_set(theme_minimal())
-theme_update(
-  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
+bb1 <- st_bbox(study_area)
+g31 <- ggplot() +
+  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
+  geom_sf(data = study_area, fill = "#666666", colour = NA, lwd = 0) +
+  annotate("rect", xmin = bb1[1], ymin = bb1[2], 
+                   xmax = bb1[3], ymax = bb1[4], fill = "#00000000", colour = "red") +
+  theme(panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
         panel.grid = element_line(colour = "black", linewidth = 0.05),
-  panel.ontop = TRUE
-)
-bb1 <- st_bbox(study_area)
-ggplot() +
-  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
-  geom_sf(data = study_area, fill = "#666666", colour = "white") +
-  annotate("rect", xmin = bb1[1], ymin = bb1[2], 
-                   xmax = bb1[3], ymax = bb1[4], fill = "#00000000", colour = "red", lwd = 0.25)
-```
-
-```{r}
-#| label: fig-study-area-map
-#| fig-cap: "Skelton and Dry Valleys basins"
+        panel.ontop = TRUE)
+
 theme_set(theme_void())
-theme_update(
-  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5)
-)
-ggplot() +
+g32 <- ggplot() +
   geom_sf(data = coastline, fill = "#eff3ff", colour = "#00000000") +
   geom_sf(data = study_basins, fill = "gray", colour = "white") +
-  geom_sf(data = coastline, fill = "#00000000", colour = "#3182bd", lwd = 0.35) +
+  geom_sf(data = coastline, fill = "#00000000", colour = "#3182bd") +
   geom_sf(data = gps_data, colour = "red", size = 0.025) +
-  geom_sf_text(data = study_basins, aes(label = NAME)) +
+  geom_sf_text(data = study_basins, aes(label = NAME), size = 3) +
   coord_sf(xlim = bb1[c(1, 3)], ylim = bb1[c(2, 4)]) +
   annotation_scale(height = unit(0.1, "cm"), location = "tr") +
-  annotation_north_arrow(style = north_arrow_minimal, which_north = "true", location = "br")
-```
-
-```{r}
-#| label: fig-dem-and-geology
-#| fig-cap: "Study area elevation (hillshade) and surface geology"
-theme_set(theme_void())
-theme_update(
-  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5),
-  legend.position = c(0.02, 0.02),
-  legend.justification = c(0, 0)
-)
+  theme(panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5))
+
 top5 <- contiguous_geologies |> 
@@
 bb2 <- st_bbox(top5)
-ggplot() +
-  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd", lwd = 0.35) +
-  geom_sf(data = study_area, fill = "#00000000", colour = "darkblue", lwd = 0.35) +
-  geom_sf(data = geology, aes(fill = POLYGTYPE), linewidth = 0) +
+g33 <- ggplot() +
+  geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
+  geom_sf(data = study_area, fill = "#00000000", colour = "darkblue") +
+  geom_sf(data = geology, aes(fill = POLYGTYPE), linewidth = 0, colour = NA) +
   scale_fill_manual(values = c("dodgerblue", "#fec44f", "#993404"), name = "Geology") + 
@@
                    colour = "red", lwd = 0.25, linetype = "dashed") +
-  coord_sf(xlim = bb1[c(1, 3)], ylim = bb1[c(2, 4)])
-```
-
-```{r}
-#| label: fig-five-subregions
-#| fig-cap: "Five sub-regions of contiguous surface geology"
-theme_set(theme_void())
-theme_update(
-  panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5)
-)
-ggplot() +
+  coord_sf(xlim = bb1[c(1, 3)], ylim = bb1[c(2, 4)]) +
+  theme(panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5))
+
+small_geologies <- contiguous_geologies |> 
+  filter(area < min(top5$area))
+g34 <- ggplot() +
   geom_sf(data = coastline, fill = "#eff3ff", colour = "#3182bd") +
-  geom_sf(data = contiguous_geologies, fill = "#eeeeee", colour = "darkgrey") +
-  geom_sf(data = top5, fill = "#fcbba1", colour = "black", linewidth = 0.25) +
+  geom_sf(data = geology, aes(fill = POLYGTYPE), linewidth = 0, colour = NA) +
+  scale_fill_manual(values = c("dodgerblue", "#fec44f", "#993404"), name = "Geology") + 
+  geom_sf(data = small_geologies, fill = "#ffffffc0", colour = "darkgrey") +
+  # geom_sf(data = top5, fill = NA, colour = "black", linewidth = 0.2) +
   guides(fill = "none") +
   annotation_scale(height = unit(0.1, "cm")) +
-  coord_sf(xlim = bb2[c(1, 3)], ylim = bb2[c(2, 4)])
-```
-
-The study area.
-:::
+  coord_sf(xlim = bb2[c(1, 3)], ylim = bb2[c(2, 4)]) +
+  theme(panel.border = element_rect(fill = NA, colour = "black", linewidth = 0.5))
+```
+
+```{r}
+#| label: fig-study-area
+#| fig-cap: "The study area: (a) Study area location in Antarctica; (b) Skelton and Dry Valleys basins; (c) Study area elevation (hillshade) and surface geology; and (d) Five sub-regions of contiguous surface geology."
+
+(g31 + g32) / (g33 + g34 + plot_layout(guides = "collect")) + 
+  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
+```
 
@@
 
-This left one persistent concern: an over-representation of consecutive fixes recorded at exactly the same elevation, resulting in many fixes with estimated slopes of exactly 0, and leading to a clearly evident dip in estimated movement speeds at 0 slope (@fig-raw-gps-data).
+This left one persistent concern: an over-representation of consecutive fixes recorded at exactly the same elevation, resulting in many fixes with estimated slopes of exactly 0, and leading to a clearly evident dip in estimated movement speeds at 0 slope (@fig-gps-data(a)).
 It is likely that these fixes are associated with GPS device drift, so a it was decided to remove all fixes where estimated slope was exactly 0.
-@fig-filtered-gps-data shows the improvement in even a crudely estimated hiking function derived from local scatterplot (LOESS) smoothing.
+@fig-gps-data(b) shows the improvement in even a crudely estimated hiking function derived from local scatterplot (LOESS) smoothing.
 Note that such functions are likely overfitted and not used further in our analysis where we favour more easily parameterised functions such as those discussed in @sec-hiking-functions.
@@
   drop_na()
-```
-
-::: {#fig-gps-data layout="[[50,-1,50]]"}
-
-```{r}
-#| label: fig-raw-gps-data
-#| fig-cap: "Boxplots by slope of speed, with smoothed estimated hiking function showing a 'dip' due to over-representation of 0 slope fixes"
-ggplot(gps_geol |> filter(slope_h >= -1, slope_h <= 1.0)) +
+
+g41 <- ggplot(gps_geol |> filter(slope_h >= -1, slope_h <= 1.0)) +
   geom_boxplot(aes(x = round(slope_h, 1), y = speed_km_h, group = round(slope_h, 1)), 
@@
 ```{r}
-#| label: fig-filtered-gps-data
-#| fig-cap: "After filtering the estimated hiking function no longer has a dip."
+#| output: false
 hexes <- gps_geol |> 
@@
 
-ggplot(gps_geol_purposive) +
+g42 <- ggplot(gps_geol_purposive) +
   geom_boxplot(aes(x = round(slope_h, 1), y = speed_km_h, group = round(slope_h, 1)), 
@@
 
-GPS data and crudely estimated hiking functions before and after filtering the to remove fixes associated with non-purposive movement.
-:::
+```{r}
+#| label: fig-gps-data
+#| fig-cap: "GPS data and crudely estimated hiking functions before and after filtering the to remove fixes associated with non-purposive movement: (a) Boxplots by slope of speed, with smoothed estimated hiking function showing a 'dip' due to over-representation of 0 slope fixes; and (b) After filtering the estimated hiking function no longer has a dip."
+g41 + g42 + 
+  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
+```
 
@@
 An important decision in constructing graphs is choice of the spacing of the hexagonal lattice, and also of the underlying DEM from which graph vertex elevations are derived.
-Given the extent of the study sites (see @fig-five-subregions) it was decided that a hexagonal lattice (see @fig-lattice-adjacencies) with hexagons equivalent in area to 100 metre square cells as appropriate.
+Given the extent of the study sites (see @fig-study-area(d)) it was decided that a hexagonal lattice (see @fig-lattice-adjacencies) with hexagons equivalent in area to 100 metre square cells as appropriate.
 The hexagon centre to centre spacing of this lattice is given by 
@@
 This network includes 30,697 vertices and 174,798 directed edges. 
-The largest of the five study sites (see @fig-five-subregions) results in a network containing almost a quarter of a million vertices and over 1.4 million directed edges.
+The largest of the five study sites (see @fig-study-area(d)) results in a network containing almost a quarter of a million vertices and over 1.4 million directed edges.
 A hiking network can be used to explore many connectivity properties of the environment.
-For example, for a chosen origin point, a _shortest path tree_ can be derived showing the route (@fig-shortest-path-tree).
+For example, for a chosen origin point, a _shortest path tree_ can be derived showing the route (@fig-hiking-network-example(c)).
 
@@
 
-::: {#fig-hiking-network-example layout="[[50,-1,50], [50,-1,50]]"}
-
-```{r}
-#| label: fig-map-of-hiking-network
-#| fig-cap: "Map of a hiking network coloured by estimated traversal times of edges" 
+```{r}
+#| output: false
 bb <- (extent |> st_bbox() |> as.vector()) + c(1.05e4, 1.25e4, -1.45e4, -1.35e4)
-ggplot() +
+g71 <- ggplot() +
   geom_raster(data = shade, aes(x = x, y = y, fill = hillshade), alpha = 0.65) +
@@
   theme_void()
-```
-
-```{r}
-#| label: fig-zoomed-in-map-of-hiking-network
-#| fig-cap: "Red-outlined area zoomed in revealing edges removed from graph due to steep slopes" 
-ggplot() +
+
+g72 <- ggplot() +
   geom_raster(data = shade, aes(x = x, y = y, fill = hillshade), alpha = 0.65) +
@@
   theme_void()
-```
-
-```{r}
-#| label: fig-shortest-path-tree
-#| fig-cap: "Shortest path tree of a hiking network from the indicated origin" 
+
 centroid <- extent |>
@@
 bb <- (extent |> st_bbox() |> as.vector()) +  + c(1.35e4, 0.8e4, -0.9e4, -1.55e4)
-ggplot() +
+g73 <- ggplot() +
   geom_sf(data = extent, colour = NA) +
@@
   theme_void()
-```
-
-```{r}
-#| label: fig-zoomed-in-shortest-path-tree
-#| fig-cap: "Red-outlined area zoomed in view of the shortest path tree" 
-ggplot() +
+
+g74 <- ggplot() +
   geom_raster(data = shade, aes(x = x, y = y, fill = hillshade), alpha = 0.65) +
@@
 
-A hiking network derived for the fifth largest study site shown in @fig-five-subregions.
-
-:::
+```{r}
+#| label: fig-hiking-network-example
+#| fig-cap: "A hiking network derived for the fifth largest study site shown in @fig-study-area(d): (a) Map of a hiking network coloured by estimated traversal times of edges; (b) Red-outlined area zoomed in revealing edges removed from graph due to steep slopes; (c) Shortest path tree of a hiking network from the indicated origin; and (d) Red-outlined area zoomed in view of the shortest path tree."
+
+(g71 + g72 + plot_layout(guides = "collect")) / (g73 + g74) + 
+  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")")
+```
 

### 2025-06-07 paper/paper.qmd
updated gitignore

@@
     self-contained: true
-  # pdf:
-  #   documentclass: article
-  #   classoption: [11pt]
-  #   toc: false
-  #   number-sections: true
-  #   number-depth: 3
-  #   linestretch: 1.15
 execute:

### 2025-06-25 paper/paper.qmd
fixed waited/weighted typo

@@
 An example result is shown in @fig-betweenness-limited-sites.
-One refinement shown in the figure is that sites have been waited by relative importance so that a site such as an expedition base camp can be expected to originate more trips than other sites.
+One refinement shown in the figure is that sites have been weighted by relative importance so that a site such as an expedition base camp can be expected to originate more trips than other sites.
 

### 2025-09-18 paper/paper.qmd
added talk to repo

@@
 ## Introduction
-**FRASER TO FRAME / CLARIFY MOTIVATION**
-
-Overview of Antarctic science: when and where, its intensity etc. Background on international treaties, etc.
-
-Relevant findings as to human impacts in Antarctica. Note that in this environment even ‘leave only footprints’ is likely impacting the environment in significant ways.
+**FRASER MATERIAL STARTS**
+
+Antarctica is often considered a vast, desolate continent devoid of human activity.
+However, there are small, scientifically significant areas where substantial, regular human activity occurs.
+Globally unique ice-free areas within Antarctica, such as the McMurdo Dry Valleys, are of particular scientific interest and so a significant amount of activity occurs in this specially managed area.
+Human activity in Antarctica is highly dependent on the infrastructure that supports people to survive in this environment. 
+Outside of a reduced period of activity caused by COVID-19, human activity on the continent continues to grow [@Tejedo2024] and so the actual, and future potential impact of these activities continues to grow.
+With the McMurdo Dry Valleys being very close to the largest research base on the continent, McMurdo Station, the pressures on this unique environment from human activity are increasing. 
+Understanding the scope and implications of human activity in key parts of Antarctica is critical to supporting the environmental management of the continent.
+
+The history of understanding the role and impacts of human activity in Antarctica begins with the sixth Antarctic Treaty Consultative Meeting (ATCM) in 1970. 
+A recommendation in the final report of this meeting was for member countries to encourage more research on how human activity has impact on Antarctica and its ecosystems [@ATCM1970 Recommendation VI-4].
+By the mid-1980s, it was recognised that scientific activity, particularly the logistics required to support that activity, required additional scrutiny, but also that scientific activity could pose a greater threat to Antarctica’s ecosystems than tourism [@benninghoff1985man].
+At this point, the need for more understanding and the ability to forecast the environmental impact of human activity was recognised [@ATCM1985 Paragraph 46].
+
+Moving into the mid-1990’s, the multi-disciplinary scope of large programmes of research like the McMurdo Dry Valleys Long Term Ecological Research (MCM LTER) programme, increased the number of researchers in the field but also supported more research into the history of human activity in the Dry Valleys [@priscu2016environmental].
+
+With the introduction of the Committee for Environmental Protection (CEP) in 1998, human activities (including science activities) were monitored, but moved away from the understanding and scope of the activities, but more focused on the potential resulting impacts of the activity.
+For example, the current 5-year Work Plan for the CEP (REF), has several items associated with human activities but are focused on managing the impact of human activities rather than quantifying the current level of activity, or forecasting the cumulative impacts of the activity.
+
+More recent research has focused either on a continental-scale picture of human activity (Leihy et al), or detailed mapping of the infrastructural footprint and impacts that the numerous Antarctic stations have on the continent (Brooks et al x 2 Papers).
+These approaches approach the challenge of understanding human activity from contrasting spatial perspectives. 
+This research aims to complement these two techniques using high resolution mapping of human activity in the dry valleys to extrapolate the way in which scientists use the landscape to move around.
+
+**FRASER MATERIAL ENDS**
 
@@
 ### GPS data from an expedition
-**FRASER**: Timeline, devices used, and associated protocols for scientists while on site.
+**FRASER MATERIAL STARTS**
+
+GPS data showing how scientists moved around the McMurdo Dry Valleys was collected using QStarz Q1100P GPS Tracking Recorders between 2016 and 2018. 
+These units were selected because of their long battery life (40+ hours), chipset sensitivity, and simple design and use.
+
+Each unit was set to record a location every 30 seconds. 
+Scientists turned on the devices when leaving the base camp in the morning and turned them off when returning at the end of the day. 
+The units were attached to the top of a backpack when leaving the camp to ensure the units had bread access to the GPS signal.
+
+Participants were sourced through existing networks of scientists supported by the New Zealand Antarctic Programme who were heading to Antarctica for research purposes. 
+All participants were asked if they wish to wear a GPS recorder. 
+The study was entirely voluntary. Over the two primary field events, 40 people used the devices. 
+Social Ethics approval was received by the Manaaki Whenua Landcare Research Social Ethics committee (SE 1617/05).
+
+![The Qstarz Q1100P GPS Tracking Recorders used to record the movement of research staff while in the McMurdo Dry Valleys.](gps-device-picture.jpg){#fig-gps-device fig-width="60.88mm"}
+
+![30 second locations recorded by ~14 researchers within a single research event within the Miers Valley.
+The image shows the basecamp, locations of research equipment and/or activities, as well as walking tracks that were taken to access areas further afield.](sample-gps-data.png){#fig-sample-gps-data fig-width="81.28mm"}
+
+**FRASER MATERIAL ENDS**
 

### 2025-09-18 talk/index.qmd
added talk

---
title: "Modelling potential environmental impacts of science activity in Antarctica"
title-slide-attributes:
    data-background-image: background-image.png
author: 
  - name: David O'Sullivan
    affiliations: 
      - name: Geospatial Stuff
      - name: University of Auckland
  - name: Fraser Morgan
    affiliations: 
      - name: Bioeconomy Science Institute
format:
  revealjs:
    theme: white
    css: custom.scss
    transition: none
    navigation-mode: vertical
    fragment-in-url: true
    slide-number: false
    logo: logo.png
    footer: "Palmerston North Regional GIS Forum, September 2025"
    self-contained: true
---

## Overview {.smaller}
### [Study area](#study-area-1)
### [GPS data](#gps-data-1)
### [Building a hiking network](#building-a-hiking-network-1)
### [Results](#results-1)
### [Conclusions](#conclusions-1)

# Study area {.smaller background-image="background-image.png" background-opacity="0.5"}

##
![](fig-study-area-1.png)

# GPS data {.smaller background-image="background-image.png" background-opacity="0.5"}

##
::: {.columns}
::: {.column width=35%}
![](gps-device-picture.jpg)
:::
::: {.column width=65%}
Collected using QStarz Q1100P GPS Tracking Recorders between 2016 and 2018

40+ hour battery life and simple use

Location recorded every 30 seconds

Scientists turned devices on when leaving base camp in the morning and off on return at the end of the day
:::
:::

## The Dry Valleys {background-image="bing-maps-imagery.png"}
[Bing maps aerial imagery]{.aside}

## And on the ground {background-image="google-street-view.jpg"}
[Yup! There's streetview imagery]{.aside}

# Hiking networks {.smaller background-image="background-image.png" background-opacity="0.5"}

## Building a hiking network

Start with a regular lattice across the study area

Many topologies could be used

```{r}
#| label: lattice-adjacencies
#| fig-width: 8
#| fig-height: 2.5

library(sf)
library(igraph)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyr)

source("raster-to-graph-functions-paper.R")

theme_set(theme_void()) 
theme_update(
  plot.margin = unit(c(1, 1, 2, 1), "mm"),
  plot.title = element_text(hjust = 0.5, vjust = 2))

get_lattice <- function(dist, pts) {
  pts |>
    st_buffer(dist) |>
    st_contains(pts, sparse = FALSE) |>
    as.matrix() |>
    graph_from_adjacency_matrix(mode = "directed", diag = FALSE) |>
    set_vertex_attr("x", value = pts$x) |>
    set_vertex_attr("y", value = pts$y)
}

get_graph_plot <- function(G, main, crs = 2193) {
  return(ggplot() +
    geom_sf(data = G |> 
              get_graph_as_line_layer(crs = 2193), lwd = 0.35) +
    geom_sf(data = G |> 
              vertex_attr() |> as.data.frame() |> st_as_sf(coords = 1:2, crs = crs), 
            size = 8, fill = "white", colour = "black", shape = 21, lwd = 1) +
    coord_sf(xlim = c(-270, 270) + 1e6, 
             ylim = c(-270, 270) + 1e6, expand = FALSE) +
    ggtitle(main))
}

grid_pts <- expand_grid(x = -20:20 * 100 + 1e6, y = -20:20 * 100 + 1e6) |>
  as.data.frame() |>
  st_as_sf(coords = c("x", "y"), crs = 2193, remove = FALSE)

hex_grid_pts <- grid_pts |>
  st_bbox() |> 
  st_make_grid(square = FALSE, what = "centers", cellsize = 107.457) |>
  st_as_sf() |>
  rename(geometry = x) |>
  mutate(geometry = geometry + c(-12, -3)) |>
  st_set_crs(2193)

xy <- hex_grid_pts |> st_coordinates()
hex_grid_pts <- hex_grid_pts |>
  mutate(x = xy[, 1], y = xy[, 2])

graphs <- mapply(
  get_lattice,
  list(101, 150, 225, 110),
  list(grid_pts, grid_pts, grid_pts, hex_grid_pts))
names(graphs) <- list("Rook's move", "Queen's move", "Knight's move", "Hex")

ggplots <- list()
for (n in names(graphs)) {
  ggplots[[n]] <- get_graph_plot(G = graphs[[n]], main = n)
}

wrap_plots(ggplots, nrow = 1)
```

We went with hexagonal at ~100m spacing

## Up $\neq$ down {background-image="klondike.jpg" background-opacity="50%"}

Slope is critical to movement, so attach heights to nodes and calculate slope in each direction along lattice edges

Then apply a *hiking function* to estimate traversal times

::: aside
[People heading uphill in the Klondike gold rush 1897](https://www.nps.gov/klgo/learn/goldrush.htm) National Park Service, Klondike Gold Rush National Historical Park, Candy Waugaman Collection, KLGO Library SS-126-8831
:::

## Hiking functions
![](fig-hiking-functions-1.png)

## Smoothing GPS data
![](fig-gps-data-1.png)

## Curve fitting to cleaned data
![](fig-comparison-of-hiking-functions-1.png)

## Terrain-dependent _vs_ combined
![](fig-models-by-cover-compared-1.png)

## Completing and applying the network

Assign edges *estimated traversal times* based on slope, the hiking function, and land cover (moraine or rock)

Make the nodes and edges into a _directed graph_

Then use various _graph algorithms_ to find, e.g., everywhere-to-everywhere shortest paths

# Some outputs {.smaller background-image="background-image.png" background-opacity="0.5"}

##
![](fig-hiking-network-example-1.png)

## Betweenness centrality
After some exploration, we settled on _betweenness centrality_ as a graph metric that might be useful

This measure counts how many times each node appears on the shortest paths between every other pair of nodes

An indicator of the relative likelihood of each location being visited

##

![](fig-betweenness-map-1.png)

## Radius-limited betweenness
Restrict betweenness centrality to nodes no more than some time apart

This is _much_ faster to calculate (yay!)

Also... it looks like it has more value

It (perhaps) is relevant to how people navigate in such environments

##

![](fig-radius-limited-betweenness-1.png)

## Planning a path network?

One way to minimise impact might be to plan paths

This is _experimental_ at this stage

Based on a _minimum spanning tree_ approximation to an _arborescence_

::: aside
A minimum spanning tree is subgraph with minimum total cost that reaches all the nodes in a network. An arborescence is the equivalent in a directed graph.
:::

##

![](fig-betweenness-limited-sites-1.png)

##

![](fig-approx-minimum-arborescence-1.png)

---

## Conclusions
Terrain differentiated fitted hiking functions are a novelty

Potential wider application of radius-limited betweenness centrality?

A manuscript almost finished

&nbsp;

#### For more
* On this project: [dosull.github.io/antarctica](https://dosull.github.io/antarctica)

* About Geospatial Stuff: [dosull.github.io](https://dosull.github.io)

# {background-image="background-image.png"}
[Questions?]{style="font-size:200px;opacity:25%;text-align:center;"}

### 2025-09-18 talk/raster-to-graph-functions-paper.R

@@ -1,0 +1,286 @@
+library(sf)
+library(terra)
+library(igraph)
+library(tidyr)
+library(dplyr)
+
+# The tobler hiking function for speed appropriately scaled to m/hr
+# if modified, use Marquez-Perez modified form which is a bit less
+# superman-ish. These formulae from movecost help. See also 
+# 
+# Tobler WR. 1993. Three Presentations on Geographical Analysis and Modeling: 
+#   Non-Isotropic Geographic Modeling; Speculations on the Geometry of Geography;
+#   and Global Spatial Analysis. National Center for Geographic Information and 
+#   Analysis, Santa Barbara, CA. https://escholarship.org/uc/item/05r820mz
+#
+tobler_speed_m_hr <- function(rise_over_run, modified = TRUE) {
+  if (modified) {
+    # Tobler off-path
+    3600 * exp(-3.5 * abs(rise_over_run + 0.05))
+  } else {
+    # Tobler - superman!
+    6 * exp(-3.5 * abs(rise_over_run + 0.05)) * 1000
+  }
+}
+
+est_hiking_function <- function(slope, terrain1, terrain2) {
+  if (terrain1 == terrain2) {
+    if (terrain1 == "moraine") {
+      4169.969 * exp(-(slope + 0.05259) ^ 2 / 0.235957)
+    } else {
+      3763.969 * exp(-(slope + 0.1188) ^ 2 / 0.364931)
+    }
+  } else {
+    (4169.969 * exp(-(slope + 0.05259) ^ 2 / 0.235957) + 3763.969 * exp(-(slope + 0.1188) ^ 2 / 0.364931)) / 2
+  }
+}
+
+est_hiking_function_vec <- Vectorize(est_hiking_function)
+
+est_hiking_function_z <- function(slope, terrain1, terrain2) {
+  if (terrain1 == terrain2) {
+    if (terrain1 == "moraine") {
+      4235.085 * exp(-(slope + 0.061365) ^ 2 / 0.1266212)
+    } else {
+      3870.854 * exp(-(slope + 0.118810) ^ 2 / 0.2975753)
+    }
+  } else {
+    (4235.085 * exp(-(slope + 0.061365) ^ 2 / 0.1266212) + 3870.854 * exp(-(slope + 0.118810) ^ 2 / 0.2975753)) / 2
+  }
+}
+
+est_hiking_function_z_vec <- Vectorize(est_hiking_function_z)
+
[…]

### 2025-09-18 talk/index.qmd
changed talk to dark theme

@@
     data-background-image: background-image.png
+    data-background-opacity: "0.5"
 author: 
@@
   revealjs:
-    theme: white
+    theme: black
     css: custom.scss
@@
 ```{r}
-#| label: lattice-adjacencies
-#| fig-width: 8
-#| fig-height: 2.5
-
 library(sf)
@@
     geom_sf(data = G |> 
-              get_graph_as_line_layer(crs = 2193), lwd = 0.35) +
+              get_graph_as_line_layer(crs = 2193), lwd = 0.35, colour = "white") +
     geom_sf(data = G |> 
               vertex_attr() |> as.data.frame() |> st_as_sf(coords = 1:2, crs = crs), 
-            size = 8, fill = "white", colour = "black", shape = 21, lwd = 1) +
+            size = 8, fill = "black", colour = "white", shape = 21, lwd = 1) +
     coord_sf(xlim = c(-270, 270) + 1e6, 
              ylim = c(-270, 270) + 1e6, expand = FALSE) +
-    ggtitle(main))
+    ggtitle(main)) +
+    theme(
+      plot.background = element_rect(fill = "black"),
+    )
 }
@@
   get_lattice,
-  list(101, 150, 225, 110),
-  list(grid_pts, grid_pts, grid_pts, hex_grid_pts))
-names(graphs) <- list("Rook's move", "Queen's move", "Knight's move", "Hex")
+  # list(101, 150, 225, 110),
+  list(101, 150, 110),
+  # list(grid_pts, grid_pts, grid_pts, hex_grid_pts))
+  list(grid_pts, grid_pts, hex_grid_pts))
+# names(graphs) <- list("Rook's move", "Queen's move", "Knight's move", "Hex")
+names(graphs) <- list("Rook's move", "Queen's move", "Hex")
+```
 
+```{r}
+#| fig-width: 8
+#| fig-height: 3.045
 ggplots <- list()
@@
 
-wrap_plots(ggplots, nrow = 1)
+wrap_plots(ggplots, nrow = 1) & theme(
+  plot.background = element_rect(fill = "black"),
+  text = element_text(colour = "white")
+)
 ```
@@
 ## Conclusions
-Terrain differentiated fitted hiking functions are a novelty
+Terrain-differentiated data-fitted hiking functions are a novelty
 
@@
 # {background-image="background-image.png"}
-[Questions?]{style="font-size:200px;opacity:25%;text-align:center;"}
+[Questions?]{style="font-size:200px;color:black;opacity:25%;text-align:center;"}

### 2025-09-24 talk/index.qmd
many minor tweaks to slides

@@
     data-background-image: background-image.png
-    data-background-opacity: "0.5"
+    data-background-opacity: "0.65"
 author: 
@@
   revealjs:
-    theme: black
-    css: custom.scss
+    theme: [black, custom.scss]
     transition: none
@@
 
-## Overview {.smaller}
-### [Study area](#study-area-1)
-### [GPS data](#gps-data-1)
-### [Building a hiking network](#building-a-hiking-network-1)
-### [Results](#results-1)
+## Overview {.smaller background-image="miers-valley.jpg"}
+### [Study area and data](#study-area-and-data-1)
+### [Hiking networks](#hiking-networks-1)
+### [Some outputs](#some-outputs-1)
 ### [Conclusions](#conclusions-1)
 
-# Study area {.smaller background-image="background-image.png" background-opacity="0.5"}
+::: aside
+Image: Miers Valley, from [commons.wikimedia.org](https://commons.wikimedia.org/wiki/File:Miers_Valley_CKL.jpg)
+:::
+
+# Study area and data {.smaller background-image="background-image.png" background-opacity="0.65"}
 
@@
 
-# GPS data {.smaller background-image="background-image.png" background-opacity="0.5"}
-
-##
+## GPS data
 ::: {.columns}
@@
 
-# Hiking networks {.smaller background-image="background-image.png" background-opacity="0.5"}
+# Hiking networks {.smaller background-image="background-image.png" background-opacity="0.65"}
 
 ## Building a hiking network
-
-Start with a regular lattice across the study area
-
-Many topologies could be used
+### Start with regular lattice across study area
+### Many topologies could be used
 
@@
 
-We went with hexagonal at ~100m spacing
+### We went with hexagonal at ~100m spacing
 
-## Up $\neq$ down {background-image="klondike.jpg" background-opacity="50%"}
-
-Slope is critical to movement, so attach heights to nodes and calculate slope in each direction along lattice edges
-
-Then apply a *hiking function* to estimate traversal times
+## Up $\neq$ down {background-image="klondike.jpg" background-opacity="0.65"}
+### Slope is critical to movement, so attach heights to nodes and calculate slope in each direction along lattice edges
+### Then apply a *hiking function* to estimate traversal times
 
@@
 ## Completing and applying the network
+### Assign edges *estimated traversal times* based on slope, the hiking function, and land cover (moraine or rock)
+### Make the nodes and edges into a _directed graph_
+### Then use various _graph algorithms_ to find, e.g., everywhere-to-everywhere shortest paths
 
-Assign edges *estimated traversal times* based on slope, the hiking function, and land cover (moraine or rock)
-
-Make the nodes and edges into a _directed graph_
-
-Then use various _graph algorithms_ to find, e.g., everywhere-to-everywhere shortest paths
-
-# Some outputs {.smaller background-image="background-image.png" background-opacity="0.5"}
+# Some outputs {.smaller background-image="background-image.png" background-opacity="0.65"}
 
@@
 ## Betweenness centrality
-After some exploration, we settled on _betweenness centrality_ as a graph metric that might be useful
+### After some exploration, we landed on _betweenness centrality_ as a useful metric
+### Counts how often each node appears on shortest paths between every other pair of nodes
+&nbsp;
 
-This measure counts how many times each node appears on the shortest paths between every other pair of nodes
-
-An indicator of the relative likelihood of each location being visited
+### &rarr; Indicator of relative likelihood of locations being visited
 
@@
 ## Radius-limited betweenness
-Restrict betweenness centrality to nodes no more than some time apart
-
-This is _much_ faster to calculate (yay!)
-
-Also... it looks like it has more value
-
-It (perhaps) is relevant to how people navigate in such environments
+### Restrict betweenness centrality to nodes no more than some cost (i.e., _time_) apart
+### _Much_ faster to calculate
+### Also... it may have more value: perhaps relevant to how people navigate in such environments
 
@@
 ## Planning a path network?
-
-One way to minimise impact might be to plan paths
-
-This is _experimental_ at this stage
-
-Based on a _minimum spanning tree_ approximation to an _arborescence_
+### A way to reduce impact might be to plan paths
+### _Experimental_ at this stage
+### Based on a _minimum spanning tree_ approximation to an _arborescence_
 
 ::: aside
-A minimum spanning tree is subgraph with minimum total cost that reaches all the nodes in a network. An arborescence is the equivalent in a directed graph.
+A minimum spanning tree is a subgraph with minimum total cost that reaches all the nodes in a network. An arborescence is the equivalent in a directed graph.
 :::
@@
 
-## Conclusions
-Terrain-differentiated data-fitted hiking functions are a novelty
-
-Potential wider application of radius-limited betweenness centrality?
-
-A manuscript almost finished
+## [Conclusions]{style="text-shadow:0px 0px 8px rgba(102,0,0,0.5);"} {background-image="beacon-valley.jpg" style="text-shadow:0px 0px 8px rgba(102,0,0,0.5);"}
+### Terrain-differentiated data-fitted hiking functions are a novelty
+### Potential wider application of radius-limited betweenness centrality?
 
@@
 #### For more
-* On this project: [dosull.github.io/antarctica](https://dosull.github.io/antarctica)
+* On this project: [[dosull.github.io/antarctica](https://dosull.github.io/antarctica)]{style="background-color:rgba(102,0,0,0.65);"}
 
-* About Geospatial Stuff: [dosull.github.io](https://dosull.github.io)
+* About ![](logo.png){style="vertical-align:middle;width:60px"} [Geospatial Stuff: [dosull.github.io](https://dosull.github.io)]{style="background-color:rgba(102,0,0,0.65);"}
+
+::: aside
+Image: Beacon Valley, from [reddit.com/rHumanForScale](https://www.reddit.com/r/HumanForScale/comments/by9q7f/beacon_valley_antarctica/)
+:::
 

### 2025-09-24 talk/index.qmd
final edits to talk
some final tweaks to slides

@@
 ### Potential wider application of radius-limited betweenness centrality?
-
-&nbsp;
-
-#### For more
-* On this project: [[dosull.github.io/antarctica](https://dosull.github.io/antarctica)]{style="background-color:rgba(102,0,0,0.65);"}
-
-* About ![](logo.png){style="vertical-align:middle;width:60px"} [Geospatial Stuff: [dosull.github.io](https://dosull.github.io)]{style="background-color:rgba(102,0,0,0.65);"}
+### Manuscript in preparation
 
@@
 # {background-image="background-image.png"}
-[Questions?]{style="font-size:200px;color:black;opacity:25%;text-align:center;"}
+
+### Questions? {style="font-size:200px;color:black;opacity:25%;text-align:center;font-weight:700;"}
+
+::: {style="background-color:rgba(102,0,0,0.1);"}
+<h2 style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);">For more</h2>
+### On this project: [dosull.github.io/antarctica](https://dosull.github.io/antarctica) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
+### About ![](logo.png){style="vertical-align:middle;width:60px;text-align:center;"} Geospatial Stuff: [dosull.github.io](https://dosull.github.io) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
+:::

### 2025-09-24 talk/index.qmd
some more minor edits to slides

--- 2025-09-24 talk/index.qmd
@@
 ##
-![](fig-study-area-1.png)
+![](fig-study-area-1.png){fig-align="center"}
 
@@
 ## Hiking functions
-![](fig-hiking-functions-1.png)
+![](fig-hiking-functions-1.png){fig-align="center"}
 
 ## Smoothing GPS data
-![](fig-gps-data-1.png)
+![](fig-gps-data-1.png){fig-align="center"}
 
 ## Curve fitting to cleaned data
-![](fig-comparison-of-hiking-functions-1.png)
+![](fig-comparison-of-hiking-functions-1.png){fig-align="center"}
 
 ## Terrain-dependent _vs_ combined
-![](fig-models-by-cover-compared-1.png)
+![](fig-models-by-cover-compared-1.png){fig-align="center"}
 
@@
 ##
-![](fig-hiking-network-example-1.png)
+![](fig-hiking-network-example-1.png){fig-align="center"}
 
@@
 
-![](fig-betweenness-map-1.png)
+![](fig-betweenness-map-1.png){fig-align="center"}
 
@@
 
-![](fig-radius-limited-betweenness-1.png)
+![](fig-radius-limited-betweenness-1.png){fig-align="center"}
 
@@
 
-![](fig-betweenness-limited-sites-1.png)
+![](fig-betweenness-limited-sites-1.png){fig-align="center"}
 
@@
 
-![](fig-approx-minimum-arborescence-1.png)
+![](fig-approx-minimum-arborescence-1.png){fig-align="center"}
 
@@
 # {background-image="background-image.png"}
-[Questions?]{style="font-size:200px;color:black;opacity:25%;text-align:center;"}
+
+### Questions? {style="font-size:220px;color:black;opacity:25%;text-align:center;font-weight:700;"}
+
+::: {style="background-color:rgba(102,0,0,0.1);"}
+<h2 style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);">For more</h2>
+### On this project [dosull.github.io/antarctica](https://dosull.github.io/antarctica) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
+### About ![](logo.png){style="vertical-align:middle;width:60px;text-align:center;"} Geospatial Stuff [dosull.github.io](https://dosull.github.io) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
+:::

### 2025-09-24 talk/index.qmd
added north arrow joke...
more fiddling around

@@
 ##
-![](fig-study-area-1.png){fig-align="center"}
+![](fig-study-area-1.png){width="80%" fig-align="center"}
+
+![](north-arrows.png){.absolute top=65 left=177 width="200" .fragment .fade-in-then-out}
+
+![](fig-study-area-1.png){.absolute top=0 left=0 width="80%" fig-align="center" .fragment}
 
@@
   return(ggplot() +
-    geom_sf(data = G |> 
-              get_graph_as_line_layer(crs = 2193), lwd = 0.35, colour = "white") +
-    geom_sf(data = G |> 
-              vertex_attr() |> as.data.frame() |> st_as_sf(coords = 1:2, crs = crs), 
-            size = 8, fill = "black", colour = "white", shape = 21, lwd = 1) +
+    geom_sf(data = G |> get_graph_as_line_layer(crs = 2193),
+            lwd = 1, colour = "grey") +
+    geom_sf(data = G |> vertex_attr() |> as.data.frame() |> 
+              st_as_sf(coords = 1:2, crs = crs), size = 12,
+              colour = "#191919", fill = "grey", shape = 21) +
     coord_sf(xlim = c(-270, 270) + 1e6, 
@@
     theme(
-      plot.background = element_rect(fill = "black"),
+      plot.background = element_rect(fill = "#191919"),
     )
@@
   rename(geometry = x) |>
-  mutate(geometry = geometry + c(-12, -3)) |>
   st_set_crs(2193)
@@
   get_lattice,
-  # list(101, 150, 225, 110),
   list(101, 150, 110),
-  # list(grid_pts, grid_pts, grid_pts, hex_grid_pts))
   list(grid_pts, grid_pts, hex_grid_pts))
-# names(graphs) <- list("Rook's move", "Queen's move", "Knight's move", "Hex")
 names(graphs) <- list("Rook's move", "Queen's move", "Hex")
@@
 wrap_plots(ggplots, nrow = 1) & theme(
-  plot.background = element_rect(fill = "black"),
+  plot.background = element_rect(fill = "#191919"),
   text = element_text(colour = "white")
@@
 
-### We went with hexagonal at ~100m spacing
+### We went with hexagonal at ~107m spacing
 
@@
 
+##
+
+![](betweenness.png){fig-align="center"}
+
 ## Planning a path network?

### 2025-09-24 talk/index.qmd
and finally finally

@@
 ## [Conclusions]{style="text-shadow:0px 0px 8px rgba(102,0,0,0.5);"} {background-image="beacon-valley.jpg" style="text-shadow:0px 0px 8px rgba(102,0,0,0.5);"}
+
+::: {style="background-color:rgba(102,0,0,0.25);"}
 ### Terrain-differentiated data-fitted hiking functions are a novelty
@@
 ### Manuscript in preparation
+### Start a conversation with Antarctic scientists about planning
+:::
 
@@
 
-::: {style="background-color:rgba(102,0,0,0.1);"}
+::: {style="background-color:rgba(102,0,0,0.2);"}
 <h2 style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);">For more</h2>
 ### On this project [dosull.github.io/antarctica](https://dosull.github.io/antarctica) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
-### About ![](logo.png){style="vertical-align:middle;width:60px;text-align:center;"} Geospatial Stuff [dosull.github.io](https://dosull.github.io) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
+### On ![](logo.png){style="vertical-align:middle;width:60px;text-align:center;"} Geospatial Stuff [dosull.github.io](https://dosull.github.io) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
 :::

### 2025-09-24 talk/index.qmd
final version of talk - made html non-self-contained

--- 2025-09-24 talk/index.qmd
@@
     footer: "Palmerston North Regional GIS Forum, September 2025"
-    self-contained: true
+    self-contained: false
 ---
@@
 ### Slope is critical to movement, so attach heights to nodes and calculate slope in each direction along lattice edges
-### Then apply a *hiking function* to estimate traversal times
+### Then use a *hiking function* to estimate traversal times
 
@@
 ## Completing and applying the network
-### Assign edges *estimated traversal times* based on slope, the hiking function, and land cover (moraine or rock)
-### Make the nodes and edges into a _directed graph_
-### Then use various _graph algorithms_ to find, e.g., everywhere-to-everywhere shortest paths
+### Estimate edge *traversal times* based on slope, hiking function, and land cover (moraine or rock)
+### Make nodes and edges into a _directed graph_
+### Use various _graph algorithms_ to find, e.g., everywhere-to-everywhere shortest paths
 
@@
 ##
-![](fig-hiking-network-example-1.png){fig-align="center"}
+![](hiking-network-example.png){fig-align="center"}
+
+##
+![](single-source-shortest-paths.png){fig-align="center"}
+
+##
+![](single-source-shortest-paths-2.png){fig-align="center"}
 
@@
 ### After some exploration, we landed on _betweenness centrality_ as a useful metric
-### Counts how often each node appears on shortest paths between every other pair of nodes
+### Counts how often each node appears on shortest paths between every other pair of nodes {.fragment}
 &nbsp;
 
-### &rarr; Indicator of relative likelihood of locations being visited
+### &rarr; Indicator of relative likelihood of locations being visited {.fragment}
 
@@
 
----
+# Conclusions {.smaller background-image="beacon-valley.jpg"}
 
-## [Conclusions]{style="text-shadow:0px 0px 8px rgba(102,0,0,0.5);"} {background-image="beacon-valley.jpg" style="text-shadow:0px 0px 8px rgba(102,0,0,0.5);"}
-### Terrain-differentiated data-fitted hiking functions are a novelty
-### Potential wider application of radius-limited betweenness centrality?
-### Manuscript in preparation
+[Image: Beacon Valley, from [reddit.com/rHumanForScale](https://www.reddit.com/r/HumanForScale/comments/by9q7f/beacon_valley_antarctica/)]{.aside .position bottom=0 left=0}
 
-::: aside
-Image: Beacon Valley, from [reddit.com/rHumanForScale](https://www.reddit.com/r/HumanForScale/comments/by9q7f/beacon_valley_antarctica/)
+## {background-image="beacon-valley.jpg"}
+
+::: {style="text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);background-color:rgba(102,0,0,0.1);backdrop-filter:blur(2px);padding:5px;border-radius:5px;"}
+### Hiking network concept is implicit in hiking functions but a new terrain representation
+### Terrain-differentiated GPS data-fitted hiking functions are novel {.fragment}
+### Potential wider application of radius-limited betweenness centrality {.fragment}
+### Next up: start a conversation with Antarctic scientists about planning {.fragment}
 :::
@@
 
-::: {style="background-color:rgba(102,0,0,0.1);"}
+::: {style="background-color:rgba(102,0,0,0.1);backdrop-filter:blur(2px);padding:5px;border-radius:5px;"}
 <h2 style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);">For more</h2>
 ### On this project [dosull.github.io/antarctica](https://dosull.github.io/antarctica) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
-### About ![](logo.png){style="vertical-align:middle;width:60px;text-align:center;"} Geospatial Stuff [dosull.github.io](https://dosull.github.io) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
+### On ![](logo.png){style="vertical-align:middle;width:60px;text-align:center;"} Geospatial Stuff [dosull.github.io](https://dosull.github.io) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
 :::

### 2025-10-08 talk/index-2.qmd
added light colour themed version

--- 2025-09-24 talk/index.qmd
@@
   revealjs:
-    theme: [black, custom.scss]
+    theme: [white, custom.scss]
     transition: none
@@
 ## Overview {.smaller background-image="miers-valley.jpg"}
+::: {style="background-color:rgba(0,51,102,0.5);border-radius:5px;padding:5px;width:42%;text-shadow:0px 0px 3px white;"}
 ### [Study area and data](#study-area-and-data-1)
@@
 ### [Conclusions](#conclusions-1)
+:::
 
@@
 ##
-![](fig-study-area-1.png){width="80%" fig-align="center"}
+![](fig-study-area-2.png){width="80%" fig-align="center"}
 
-![](north-arrows.png){.absolute top=65 left=177 width="200" .fragment .fade-in-then-out}
+![](north-arrows.png){.absolute top=65 left=179 width="200" .fragment .fade-in-then-out}
 
@@
     geom_sf(data = G |> get_graph_as_line_layer(crs = 2193),
-            lwd = 1, colour = "grey") +
+            lwd = 1, colour = "darkgrey") +
     geom_sf(data = G |> vertex_attr() |> as.data.frame() |> 
               st_as_sf(coords = 1:2, crs = crs), size = 12,
-              colour = "#191919", fill = "grey", shape = 21) +
+              colour = "white", fill = "darkgrey", shape = 21) +
     coord_sf(xlim = c(-270, 270) + 1e6, 
              ylim = c(-270, 270) + 1e6, expand = FALSE) +
-    ggtitle(main)) +
+    ggtitle(main) +
     theme(
-      plot.background = element_rect(fill = "#191919"),
-    )
+      plot.background = element_rect(fill = "white"),
+    ))
 }
@@
 wrap_plots(ggplots, nrow = 1) & theme(
-  plot.background = element_rect(fill = "#191919"),
-  text = element_text(colour = "white")
+  plot.background = element_rect(fill = "white", colour = NA),
+  text = element_text(colour = "black")
 )
@@
 
-## Up $\neq$ down {background-image="klondike.jpg" background-opacity="0.65"}
+## Up &#8800; down {background-image="klondike.jpg" background-opacity="0.5"}
 ### Slope is critical to movement, so attach heights to nodes and calculate slope in each direction along lattice edges
@@
 
-![](fig-betweenness-map-1.png){fig-align="center"}
+![](fig-betweenness-map-2.png){fig-align="center"}
 
@@
 
-![](fig-radius-limited-betweenness-1.png){fig-align="center"}
+![](fig-radius-limited-betweenness-2.png){fig-align="center"}
 
@@
 
-![](fig-betweenness-limited-sites-1.png){fig-align="center"}
+![](fig-betweenness-limited-sites-2.png){fig-align="center"}
 
@@
 
-![](fig-approx-minimum-arborescence-1.png){fig-align="center"}
+![](fig-approx-minimum-arborescence-2.png){fig-align="center"}
 
-# Conclusions {.smaller background-image="beacon-valley.jpg"}
+# [Conclusions]{style="color:white;"} {.smaller background-image="beacon-valley.jpg" style="color:white;"}
 
@@
 
-::: {style="text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);background-color:rgba(102,0,0,0.1);backdrop-filter:blur(2px);padding:5px;border-radius:5px;"}
-### Hiking network concept is implicit in hiking functions but a new terrain representation
-### Terrain-differentiated GPS data-fitted hiking functions are novel {.fragment}
-### Potential wider application of radius-limited betweenness centrality {.fragment}
-### Next up: start a conversation with Antarctic scientists about planning {.fragment}
+::: {style="text-shadow:0px 0px 8px rgba(102,0,0,0.5),0px 0px 4px rgba(102,0,0,0.65);background-color:rgba(102,0,0,0.1);backdrop-filter:blur(2px);padding:5px;border-radius:5px;"}
+### Hiking network concept is implicit in hiking functions but a new terrain representation {style="color:white;"}
+### Terrain-differentiated GPS data-fitted hiking functions are novel {.fragment style="color:white;"}
+### Potential wider application of radius-limited betweenness centrality {.fragment style="color:white;"}
+### Next up: start a conversation with Antarctic scientists about planning {.fragment style="color:white;"}
 :::
@@
 ::: {style="background-color:rgba(102,0,0,0.1);backdrop-filter:blur(2px);padding:5px;border-radius:5px;"}
-<h2 style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);">For more</h2>
-### On this project [dosull.github.io/antarctica](https://dosull.github.io/antarctica) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
-### On ![](logo.png){style="vertical-align:middle;width:60px;text-align:center;"} Geospatial Stuff [dosull.github.io](https://dosull.github.io) {style="text-align:center;text-shadow:0px 0px 8px rgba(102,0,0,0.35),0px 0px 4px rgba(102,0,0,0.65);"}
+<h2 style="text-align:center;">For more</h2>
+### On this project [dosull.github.io/antarctica](https://dosull.github.io/antarctica) {style="text-align:center;"}
+### On ![](logo.png){style="vertical-align:middle;width:60px;text-align:center;"} Geospatial Stuff [dosull.github.io](https://dosull.github.io) {style="text-align:center;"}
 :::

### 2025-10-09 paper/paper.qmd
latest iteration of paper - discussion and conclusions still to come

--- 2025-06-25 paper/paper.qmd
@@
 ## Introduction
-**FRASER TO FRAME / CLARIFY MOTIVATION**
-
-Overview of Antarctic science: when and where, its intensity etc. Background on international treaties, etc.
-
-Relevant findings as to human impacts in Antarctica. Note that in this environment even ‘leave only footprints’ is likely impacting the environment in significant ways.
-
-Overview of sections ahead.
+**FRASER MATERIAL STARTS**
+
+Antarctica is often considered a vast, desolate continent devoid of human activity.
+However, there are small, scientifically significant areas where substantial, regular human activity occurs.
+Globally unique ice-free areas within Antarctica, such as the McMurdo Dry Valleys, are of particular scientific interest and so a significant amount of activity occurs in this specially managed area.
+Human activity in Antarctica is highly dependent on the infrastructure that supports people to survive in this environment. 
+Outside of a reduced period of activity caused by COVID-19, human activity on the continent continues to grow [@Tejedo2024] and so the actual, and future potential impact of these activities continues to grow.
+The McMurdo Dry Valleys are close to the largest research base on the continent, McMurdo Station, and the pressures on this unique environment from human activity are increasing. 
+Understanding the scope and implications of human activity in key parts of Antarctica is critical to supporting environmental management of the continent.
+
+The history of understanding the role and impacts of human activity in Antarctica starts with the Sixth Antarctic Treaty Consultative Meeting (ATCM) in 1970. 
+A recommendation of the final report of this meeting was for member countries to encourage research on how human activity impacts Antarctica and its ecosystems [@ATCM1970 Recommendation VI-4].
+By the mid-1980s, it was recognised that scientific activity, particularly the logistics required to support that activity, required additional scrutiny, and also that scientific activity could pose a greater threat to Antarctica’s ecosystems than tourism [@Benninghoff1985].
+At this point, the need for more understanding and the ability to forecast the environmental impact of human activity was recognised [@ATCM1985 Paragraph 46].
+
+Moving into the mid-1990s, the multi-disciplinary scope of large programmes of research such as the McMurdo Dry Valleys Long Term Ecological Research (MCM LTER) programme, increased the number of researchers in the field while also supporting more research into the history of human activity in the Dry Valleys [@Priscu2016].
+With the establishment of the Committee for Environmental Protection (CEP) in 1998, human activities (including science activities) were to be monitored, but attention moved away from the understanding and scope of the activities, and became more focused on potential resulting impacts of those activities.
+For example, the current 5-year Work Plan for the CEP (REF), has several items associated with human activities but these are focused on managing the impact of human activities rather than quantifying the current level of activity, or forecasting cumulative impacts of the activity.
+
+More recent research has focused either on a continental-scale picture of human activity [@Leihy2020], or detailed mapping of the infrastructural footprint and impacts that the numerous Antarctic stations have on the continent [@Brooks2018,@Brooks2019].
+These approaches tackle the challenge of understanding human activity from contrasting spatial perspectives. 
+The present research aims to complement these two techniques, by using high resolution mapping of human activity in the dry valleys to extrapolate how scientists might navigate around these landscapes.
+
+In the next section we outline our approach and discuss related work.
+We then describe data sources, both general environmental and topographic, and the detailed GPS data from which our findings are derived.
+Procedures for deriving hiking functions and applying them to build hiking networks as representations of the dry valleys landscapes are explained, followed by examples of analyses possible using this approach, particularly the application of measures of graph betweenness centrality as proxies for potential impact.
+We also outline by which these methods might be used to plan for reduced impact from scientific expeditions to these environments.
+The paper concludes with a discussion of our findings, prospects for further work, and some general conclusions.
 
@@
 @Prisner2023 provide an overview of a variety of functions that have been used to model how hiking times and speeds vary with terrain slope. 
-They consider longstanding rules of thumb [@Naismith1892], and later modifications thereof [@Langmuir1984], along with more recent such guidance from the Swiss and German Alpine Clubs [@Winkler2010; @Deutscher2011].
+They consider longstanding rules of thumb [@Naismith1892], and later modifications thereof [@Langmuir1984], along with more recent such guidance from the Swiss and German Alpine Clubs [@Winkler2010; @DeutscherundOstereichischerAlpenverein2011].
 These functions estimate the time taken to travel 1km, referred to as *pace*, based on slope expressed as *rise over run*, that is change in elevation divided by horizontal distance. 
@@
 
-Alongside these hiking pace functions @Prisner2023 also present hiking speed functions (generally referred to as simply *hiking functions*) from Tobler [-@Tobler1993 generally considered the first hiking function] and more recent, related but more firmly empirically grounded alternatives offered by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019.
+Alongside these hiking pace functions @Prisner2023 also present hiking speed functions (generally referred to as simply *hiking functions*) from Tobler [-@Tobler1993 generally considered the first hiking function] and more recent, related but more firmly empirically grounded alternatives offered by @Marquez-Perez2017, @Irmischer2018, and @Campbell2019a.
 Another hiking function not discussed by @Prisner2023 is presented by @Rees2004. 
@@
 | Exponential | $ae^{b\left|\theta-c\right|}$            | @Tobler1993, @Marquez-Perez2017 |
-| Gaussian    | $ae^{-b(\theta-c)^2}$                    | @Irmischer2018, @Campbell2019   |
-| Lorentz     | $\frac{a}{\left[b+d(\theta-c)^2\right]}$ | @Campbell2019                   |
+| Gaussian    | $ae^{-b(\theta-c)^2}$                    | @Irmischer2018, @Campbell2019a   |
+| Lorentz     | $\frac{a}{\left[b+d(\theta-c)^2\right]}$ | @Campbell2019a                   |
 | Quadratic   | $a+b\theta+c\theta^2$                    | @Rees2004                       |
@@
 #| label: fig-hiking-functions
-#| fig-cap: "Example hiking functional forms: Exponential [@Tobler1993], Gaussian [@Irmischer2018], Lorentz [@Campbell2019], and Quadratic [@Rees2004]."
+#| fig-cap: "Example hiking functional forms: Exponential [@Tobler1993], Gaussian [@Irmischer2018], Lorentz [@Campbell2019a], and Quadratic [@Rees2004]."
 
@@
 The parameterisations of the functions in @fig-hiking-functions have been chosen for illustrative purposes only, although the parameter values for the exponential function shown are such that $v=6e^{-3.5\left|\theta+0.05\right|}$ as suggested in @Tobler1993.
-These parameters give *Tobler's hiking function* but, as has been noted elsewhere, [see @Herzog2010; @Campbell2019], are based on a poorly specified fit to secondary data on hiking speeds found in Imhof [-@Imhof1950, pp.
+These parameters give *Tobler's hiking function* but, as has been noted elsewhere, [see @Herzog2010; @Campbell2019a], are based on a poorly specified fit to secondary data on hiking speeds found in Imhof [-@Imhof1950, pp.
 217-220].
@@
 In general, approaches based on minimising energy use yield similar results to empirical speed-slope functions, although it is worth noting that they more reliably generate zig-zag or 'switchback' movement behaviour on steep slopes [@Llobera2007].
-However, these are hard to implement, and we have adopted empirically-derived and locally-specific hiking functions following @Marquez-Perez2017, @Irmischer2018, and @Campbell2019. 
+However, these are hard to implement, and we have adopted empirically-derived and locally-specific hiking functions following @Marquez-Perez2017, @Irmischer2018, and @Campbell2019a. 
 This choice is based on available data and the goals of our study, where the relative cost of different potential routes in the landscape is more important than exact prediction of routes. 
@@
 Such smoothing will usually result in shorter edge traversal times along such edges than those that would accumulate along intervening sections of varying slope. 
-This problem is discussed in a slightly different context by Campbell et al. [-@Campbell2019 pp. 96-98].
+This problem is discussed in a slightly different context by Campbell et al. [-@Campbell2019a pp. 96-98].
 
@@
 ### GPS data from an expedition
-**FRASER**: Timeline, devices used, and associated protocols for scientists while on site.
-
-GPS data were processed to make them better suited for use in the estimation of hiking functions.
-
+GPS data recording how scientists moved around the McMurdo Dry Valleys was collected using QStarz Q1100P GPS Tracking Recorders between 2016 and 2018 (see @fig-gps-device).
+These units were selected for their long battery life (40+ hours), chipset sensitivity, and simple design and use.
+
+![The Qstarz Q1100P GPS Tracking Recorders used to record the movement of research staff while in the McMurdo Dry Valleys.](gps-device-picture.jpg){#fig-gps-device fig-width="60.88mm"}
+
+Each unit was set to record a location every 30 seconds. 
+Scientists turned on the devices when leaving base camp in the morning and turned them off when returning at the end of the day.
+The units were attached to the top of a backpack to ensure units had good access to the GPS signal.
+
+Participants were sourced through existing networks of scientists supported by the New Zealand Antarctic Programme who were going to Antarctica for research purposes. 
+All participants were asked if they wished to carry a GPS recorder. 
+The study was entirely voluntary. 
+Over the duration of the two primary field events, 40 people used the devices. 
+Social Ethics approval was granted by the Manaaki Whenua Landcare Research Social Ethics committee (SE 1617/05).
+
+Sample output of the GPS data is shown in in @fig-sample-gps-data.
+
+![30 second locations recorded by ~14 researchers during a single research trip to the Miers Valley.
+The image shows the basecamp, locations of research equipment and/or activities, movement among these locations and to access areas further afield.](gps-traces-figure.png){#fig-sample-gps-data fig-width="100mm"}
+
+It was necessary to process GPS data to make them better suited for use in the estimation of hiking functions. 
 The first processing step was to confirm the plausibility of the data, particularly the device-generated speed distance between fixes, and elevations associated with fixes.
@@
 ### Hiking functions {#sec-hiking-function-estimation}
-We fit three alternative functional forms to the cleaned GPS data: exponential [@Tobler1993], Gaussian [following @Irmischer2018], and Lorentz [following @Campbell2019] using the Levenburg-Marquardt algorithm [@More1978] as provided by the `nlsLM` function in the `minpack.lm` R package [@Elzhov2022].
+We fit three alternative functional forms to the cleaned GPS data: exponential [@Tobler1993], Gaussian [following @Irmischer2018], and Lorentz [following @Campbell2019a] using the Levenburg-Marquardt algorithm [@More1978] as provided by the `nlsLM` function in the `minpack.lm` R package [@Elzhov2022].
 The raw data and fitted curves are shown in @fig-comparison-of-hiking-functions.
@@
 ### Landscapes as graphs
-We developed R code [@RCoreTeam2024] to build graphs (i.e. networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions. 
-Graphs are manipulated as `igraph` package [@csardi2006igraph; @Csardi2024] graph objects for further analysis.
+We developed R code [@RCoreTeam2025] to build graphs (i.e. networks) with hexagonal lattice structure and estimated traversal times for graph edges derived from our hiking functions. 
+Graphs are manipulated as `igraph` package [@Csardi2006a; @Csardi2024] graph objects for further analysis.
 An important decision in constructing graphs is choice of the spacing of the hexagonal lattice, and also of the underlying DEM from which graph vertex elevations are derived.
@@
   + Possible limitations of a regularly spaced lattice: extend to a lattice based on e.g., TIN?
-  + Reliance of shortest paths as a guide to behaviour
+  + Reliance on shortest paths as a guide to behaviour
   + Some minor issues regarding the difference between directed and undirected graphs

### 2025-12-17 paper/paper.qmd
removed suggestion of figure of largest site
latest version of paper

@@
 ## Introduction
-**FRASER MATERIAL STARTS**
-
 Antarctica is often considered a vast, desolate continent devoid of human activity.
@@
 
-More recent research has focused either on a continental-scale picture of human activity [@Leihy2020], or detailed mapping of the infrastructural footprint and impacts that the numerous Antarctic stations have on the continent [@Brooks2018,@Brooks2019].
+More recent research has focused either on a continental-scale picture of human activity [@Leihy2020], or detailed mapping of the infrastructural footprint and impacts that the numerous Antarctic stations have on the continent [@Brooks2018; @Brooks2019].
 These approaches tackle the challenge of understanding human activity from contrasting spatial perspectives. 
-The present research aims to complement these two techniques, by using high resolution mapping of human activity in the dry valleys to extrapolate how scientists might navigate around these landscapes.
+The present research aims to complement these two techniques, by using high resolution mapping of human activity in the dry valleys to extrapolate how scientists navigate in these landscapes, or could potentially navigate in these landscapes.
 
@@
 ## Discussion {#sec-discussion}
-To cover:
-
-+ Figure of the largest study site?
 + Limitations of the method:
-  + Data concerns, esp. opportunistic nature of the GPS data
-  + Would higher resolution network be useful?
+  + Data concerns, esp. given improvements in  GPS data collection since 2016/18
+  + Would higher resolution network be useful? (we think probably not)
   + Possible limitations of a regularly spaced lattice: extend to a lattice based on e.g., TIN?
   + Reliance on shortest paths as a guide to behaviour
-  + Some minor issues regarding the difference between directed and undirected graphs
+  + Some minor issues regarding the difference between directed and undirected graphs (to check using `networkx`)
 + Relevance of radius limits on betweenness: the implicit model of movement in this approach (short horizons, no existing paths, 'keep going for half an hour in this general direction...')
-+ Question of embedding paths, i.e., locking in more damage to some parts of the landscape vs others
++ Question of embedding paths, i.e., locking in more damage to some parts of the landscape vs others (this leads to the question in conclusion re method for less damaging plans)
 
